---
title: "Corridor detection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Corridor detection}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 5,
  message = TRUE
)
```

`mt_corridor()` identifies segments of a track where the animal was
moving quickly in a spatially consistent direction. It is a port of
the legacy `move::corridor()` function (LaPoint et al. 2013) to the
modern `move2` / `sf` stack — the concept is unchanged.

## When to use corridor detection

The typical context is a migratory or commuting animal that travels
between two (or several) regions via recognisable routes. Within
those routes, successive segments point the same way and are
traversed at speed; elsewhere on the track the animal forages or
rests at slow, directionally inconsistent speeds. Corridor detection
asks: *which segments of the track belong to the route, and which
belong to the foraging / resting stretches at each end?*

## The algorithm in one paragraph

For each segment *i* of the track, compute its speed and azimuth.
Build a search circle centred at the segment midpoint with radius
equal to half the segment length. Collect all other segment midpoints
inside that circle. Compute the circular variance of
**pseudo-azimuths** (`(2 * azimuth) mod 2π`, on azimuths in radians)
within the neighbourhood — the doubling is what lets two animals walking the
same corridor in opposite directions count as consistent. A segment
is flagged corridor if (a) its speed is at least `speed_threshold`,
(b) its neighbourhood's circular variance is at most
`circvar_threshold`, and (c) the neighbourhood contains at least
`min_segments` qualifying neighbours, which must outnumber the
non-qualifying ones.

## Coordinate reference system

The function works in any CRS: longitude/latitude or projected.
Longlat input is reprojected internally to a local azimuthal-equidistant
projection (via `move2::mt_aeqd_crs()`) just for the buffer-and-index
step; the result is added back to the input in its original CRS.
Multi-track input is handled track-by-track natively, via
`move2::mt_segments()` — no cross-track segments are ever created.

## Thresholds

Both `speed_threshold` (m/s) and `circvar_threshold` (dimensionless,
in `[0, 1]`) are user-supplied numbers. If you leave them as `NULL`
the function falls back to within-object quantiles (0.75 of segment
speeds; 0.25 of valid circular variances) and **warns**, naming the
resolved numeric. The warning is intentional: a within-object
quantile defines "fast" and "directionally consistent" *relative to
this animal*, which is convenient for single-track exploration but
makes corridor maps from different individuals incomparable.

For comparable corridors across individuals or populations, compute
the thresholds once on the pooled cohort and pass them as numbers.

## Single-individual exploration

```{r load, message = FALSE}
library(move2)
library(sf)
library(move2utils)

fishers <- mt_read(mt_example())
## drop empty geometries up front; mt_corridor() would otherwise emit
## a one-line inform notice and NA-pad them in the output, but several
## downstream sf/move2 calls in this vignette (e.g. mt_segments() for
## plotting) require non-empty points throughout.
fishers <- fishers[!st_is_empty(fishers), ]
leroy   <- filter_track_data(fishers, .track_id = "M1")
st_crs(leroy)$input
```

```{r leroy-corridor}
## defaults: within-object quantile fallback fires for both thresholds
out <- mt_corridor(leroy)
table(out$corridor)
```

Five columns are added to the input `move2` (each of length
`nrow(leroy)`):

- `corridor` — factor with levels `"corridor"` and `"not corridor"`.
- `corridor_speed` — per-segment speed (m/s); `NA` at the last row of
  each track and at any row that started empty.
- `corridor_azimuth` — per-segment azimuth (degrees); `NA` likewise.
- `corridor_circvar` — circular variance of pseudo-azimuths within
  the segment's neighbourhood; `NA` for neighbourhoods of size < 2.
- `corridor_n_neighbours` — number of segments found within the
  search radius.

### A quick map

```{r leroy-map, fig.height = 5.5, fig.cap = "Leroy's track with the minority of segments that meet the corridor criteria highlighted in firebrick. Most of the track — slower, tightly looping foraging movement — stays grey."}
corr <- as.character(out$corridor)

out$segments <- mt_segments(out)
plot(out$segments,
     col = c("firebrick", "grey80")[as.factor(out$corridor)],
     lwd = c(2, 0.7)[as.factor(out$corridor)],
     asp = 1,
     main = sprintf("Leroy (fisher): %d corridor segments of %d",
                    sum(corr == "corridor"), length(corr)))
points(out,
       col = c("firebrick", "grey40")[as.factor(out$corridor)],
       pch = 16,
       cex = c(0.6, 0.3)[as.factor(out$corridor)])
legend("topright", pch = 16, col = c("grey40", "firebrick"),
       legend = c("not corridor", "corridor"), bty = "n")
```

Tightening the thresholds makes the detector more conservative —
only the fastest, most directionally consistent portions qualify.

## Multi-individual: explicit thresholds for comparability

When you want comparable corridor flags across individuals, compute
the speed threshold once on the pooled cohort and reuse it. Below,
the function is fed the whole multi-track fisher dataset directly;
each track is processed independently for segments and neighbourhoods
but every track is judged against the *same* speed cut-off.

```{r multi-individual}
## one pooled speed cut-off, reused per individual
all_speeds      <- as.numeric(mt_speed(fishers, units = "m/s"))
speed_cut       <- stats::quantile(all_speeds, 0.75, na.rm = TRUE)

out_all <- mt_corridor(
  fishers,
  speed_threshold   = speed_cut,
  circvar_threshold = 0.2   # fixed value; pick from the data or theory
)
table(mt_track_id(out_all), out_all$corridor)
```

The same threshold for `speed_threshold` and `circvar_threshold` is
now applied to each individual — so the `corridor` column means the
same thing in track M1 and track M2.

## Tuning

| Argument | Default | What to tune it for |
|---|---|---|
| `speed_threshold` | `NULL` (→ within-object 0.75 quantile, warn) | Per-segment speed in m/s. Supply a number to make corridor flags comparable across animals; raise for stricter speed cut-off. |
| `circvar_threshold` | `NULL` (→ within-object 0.25 quantile, warn) | Per-neighbourhood circular variance in `[0, 1]`. Lower for tighter directional consistency; raise if the corridor genuinely winds. |
| `min_segments`     | 2  | Minimum neighbourhood support. Raise if you want only well-populated corridor stretches. |
| `verbose`          | `FALSE` | Set to `TRUE` to print a one-line summary of how many segments were flagged. The default-threshold, irregular-sampling, and empty-geometry notices are shown regardless. |

## Further reading

- LaPoint, S., Gallery, P., Wikelski, M., & Kays, R. (2013). Animal
  behavior, cost-based corridor models, and real corridors.
  *Landscape Ecology*, 28, 1615–1630.
  <https://doi.org/10.1007/s10980-013-9910-0>
- `?mt_corridor` for the full argument reference.
