---
title: "Interpolating and thinning trajectories"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interpolating and thinning trajectories}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 5,
  message = TRUE
)
```

Real tracks are sampled irregularly, and some downstream analyses assume either 
regular sampling or a specific geometry. This vignette covers:

- **Interpolation** — producing additional locations at a desired
  cadence or along a target line. Handled by `move2::mt_interpolate()`.
- **Time-thinning** — downsampling a dense track to a target cadence.
  Two complementary helpers: `move2::mt_filter_per_interval()` for
  the bucketed case, and `move2utils::mt_thin_time()` for
  tolerance-constrained downsampling.
- **Distance-thinning** — downsampling based on cumulative travel
  distance. Provided by `move2utils::mt_thin_distance()`.


## Setting the scene

```{r load, message = FALSE}
library(move2)
library(dplyr)
library(sf)
library(move2utils)  # loaded here so that future helpers are visible

fishers <- mt_read(mt_example())
fishers <- fishers[!st_is_empty(fishers), ]

leroy <- filter_track_data(fishers, .track_id = "M1")
leroy <- slice(leroy,1:300)

summary(mt_time_lags(leroy, units = "min"))
```

## Interpolation

`mt_interpolate()` supports three use patterns.

### A. Regular time cadence via an interval string

```{r interp-interval}
## one fix per five minutes, refusing to bridge gaps longer than 1 h
leroy_5min <- mt_interpolate(
  leroy,
  time         = "5 mins",
  max_time_lag = as.difftime(1, units = "hours")
)
nrow(leroy); nrow(leroy_5min)
summary(mt_time_lags(leroy_5min, units = "min"))
```

The `"5 mins"` string is passed through `lubridate::floor_date()`, so
you can use any interval format that accepts (`"30 secs"`, `"2 hours"`,
`"1 day"`, ...). `max_time_lag` is the policy for gaps: any original
lag longer than this cutoff is *not* bridged; the output track has a
break there rather than a dense string of interpolated points
across the gap.

### B. Interpolate at a specific set of timestamps

```{r interp-specific}
t0 <- min(mt_time(leroy))
target_t <- t0 + as.difftime(c(10, 20, 30, 60, 120), units = "mins")
leroy_at <- mt_interpolate(leroy, time = target_t, omit = FALSE)
## the new locations
head(leroy_at[mt_time(leroy_at) %in% target_t, ])
```

Setting `omit = TRUE` returns only the requested timestamps and
discards the originals.

### C. Interpolate to where a line crosses the track

Useful for finding when an animal crossed a boundary — a road, a
home-range edge, a territory line.

```{r interp-line, eval = FALSE}
## example: bounding square line (plug in your own geometry in practice)
line <- sf::st_sfc(
  sf::st_linestring(cbind(c(-1, 1, 1, -1, -1), c(-1, -1, 1, 1, -1))),
  crs = sf::st_crs(leroy)
)
crossings <- mt_interpolate(leroy, sf = line)
```

### Comparison with legacy `move::interpolateTime()`

| move                                               | move2                                      |
|----------------------------------------------------|--------------------------------------------|
| `interpolateTime(x, time = 500)` (count)           | `mt_interpolate(x, time = seq(min(t), max(t), length.out = 500))` |
| `interpolateTime(x, time = as.difftime(5, "mins"))`| `mt_interpolate(x, time = "5 mins", max_time_lag = ...)`          |
| `interpolateTime(x, time = timestamp_vec)`         | `mt_interpolate(x, time = timestamp_vec)` |

The explicit `max_time_lag` is slightly more disciplined than the
legacy default — you decide what counts as "too long to interpolate"
rather than letting the function silently fill every gap.

## Time-thinning

The question "give me ~one fix per 45 min" has two sensibly different
answers. `move2::mt_filter_per_interval()` gives the **bucketed**
answer: divide time into 45-minute windows (aligned to wall-clock
intervals via `floor_date`), and pick one fix per window.

```{r filter-per-interval}
leroy_45 <- mt_filter_per_interval(leroy, unit = "45 min",
                                    criterion = "first")
nrow(leroy); nrow(leroy_45)
summary(mt_time_lags(leroy_45, units = "min"))
```

The criterion can be `"first"`, `"last"`, or `"random"`. Because the
buckets are aligned to the wall clock, the actual time lag between
retained fixes is not guaranteed to equal 45 minutes; if the first
observed fix of a window falls at 10:55 and the next at 11:00, both
are kept (they belong to different hour windows but are only 5 minutes
apart in time).

### Tolerance-constrained thinning with `mt_thin_time()`

`mt_filter_per_interval()` is fine for visualisation, quick-look
downsampling, or feeding a method that is tolerant of irregular
sampling.

`mt_thin_time()` solves a stricter problem: find the largest subset
of fixes such that **every successive retained pair is within
`[interval - tolerance, interval + tolerance]`**. The answer is the
longest path in a small DAG on sorted timestamps; it is computed with
a linear-time dynamic-programming sweep rather than the memoised
recursion of the legacy implementation, so it scales cleanly to
\eqn{10^5}{1e5}+ fixes.

```{r thin-time}
thinned <- mt_thin_time(leroy,
                         interval  = "30 min",
                         tolerance = "5 min",
                         remove    = TRUE)
nrow(leroy); nrow(thinned)

lags <- mt_time_lags(thinned, units = "min")
summary(lags[!is.na(lags)])
```

The retained-lag distribution shows the two regimes the function
produces by construction:

- **Within-run lags** fall inside `[interval - tolerance,
  interval + tolerance]` — these are the tolerance-admissible
  transitions that the DP maximised.
- **Between-run lags** exceed `interval + tolerance` — they are the
  original gaps that were too long to bridge; the run-splitter pre-pass
  decomposed the track at these gaps so that thinning never pretends
  to bridge them.

No retained lag is ever shorter than `interval - tolerance`; that is
the hard contract the function enforces.

Tie-breaking when several chains are equally long:

- `criterion = "closest"` (default) — minimum summed deviation from
  the target interval, preferring chains that hit the target most
  evenly.
- `criterion = "first"` — earliest-starting chain, matching the
  legacy `"first"` behaviour.


## Distance-thinning

"Give me one fix every 300 metres of travel" is handled by
`mt_thin_distance()`. It walks along the track, accumulating
great-circle (lon/lat) or planar (projected) distance via
`move2::mt_distance()`. Like the time case, the same question has two
sensible answers, chosen with `method`.

The default, `method = "step"`, is the greedy one: keep the first fix,
then keep a fix each time the cumulative travel **since the last
retained fix** reaches `distance`. Every retained pair is then at least
`distance` of travel apart.

```{r thin-distance}
leroy_300m <- mt_thin_distance(leroy, distance = 300)
nrow(leroy); sum(leroy_300m$thin_selected)

## filtered output for downstream use
leroy_300m_kept <- mt_thin_distance(leroy, distance = 300, remove = TRUE)
```

`method = "interval"` is the distance-space twin of `mt_thin_time()`:
it finds the *largest* subset in which every successive retained pair
falls within `[distance - tolerance, distance + tolerance]` of travel,
using the same dynamic-programming sweep. Reach for it when you want a
spacing band rather than a floor.

```{r thin-distance-interval}
leroy_300pm30 <- mt_thin_distance(leroy, distance = 300, tolerance = 30,
                                  method = "interval", remove = TRUE)
nrow(leroy_300pm30)
```

In both cases multi-track input is thinned per individual and the
output is labelled in place with a `thin_selected` column (matching the
`mt_thin_time()` convention). With `method = "step"` the first fix of
each track is always retained; with `method = "interval"` the
longest-chain selection may start later. `distance` can be a numeric
metre count or a `units` object (`units::set_units(0.5, "km")`), so the
call can read naturally.

## Summary

| Task | Function |
|---|---|
| Interpolate at regular intervals | `mt_interpolate(x, time = "5 mins", max_time_lag = ...)` |
| Interpolate at specific timestamps | `mt_interpolate(x, time = t_vec)` |
| Interpolate where a line crosses | `mt_interpolate(x, sf = line)` |
| Time-thin (bucketed) | `mt_filter_per_interval(x, unit = "45 min", criterion = "first")` |
| Time-thin (tolerance-constrained) | `mt_thin_time(x, interval, tolerance)` |
| Distance-thin | `mt_thin_distance(x, distance)` |

## Further reading

- `?move2::mt_interpolate`, `?move2::mt_filter_per_interval`.
