---
title: "Cleaning a white stork GPS track: a narrated pipeline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Cleaning a white stork GPS track: a narrated pipeline}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 8,
  fig.height = 5,
  dpi = 90,
  message = FALSE,
  warning = FALSE
)
```

This vignette cleans a real, high-frequency GPS track from start to
finish: **Pettstadt1**, a juvenile white stork (*Ciconia ciconia*) from
the **LifeTrack White Stork Bavaria** study on Movebank (study ID
24442409). The full track has 49,263 fixes between June 2024 and April
2026, and it carries several kinds of error at once: huge "teleport"
GPS failures, fixes taken with too few satellites, borderline movement
oddities within Europe, and the usual download artefacts (empty
geometries and duplicate timestamps).

We clean it in four stages, and each stage removes a distinct kind of
error:

1. **Structural cleaning** — drop empty geometries and duplicate
   timestamps.
2. **GPS-quality pre-filter** — `mt_filter_gps_quality()` removes fixes
   with too few satellites, high dilution of precision (DOP), or poor
   e-obs horizontal accuracy. This is the first line of defence.
3. **Path-position check** — `mt_flag_outliers_bridge()` scores each
   fix against where its time-neighbours place it. It does not use a
   locally estimated variance, so a bad fix cannot hide itself; it
   catches the teleport-class errors that slip past the quality filter.
4. **Unusual-movement check** — `mt_flag_outliers()` scores each fix by
   how unlikely its step length, turning angle, and their gap-aware
   auto-differences are together. It catches fixes that sit in a
   plausible place but move in a way the animal never normally does.

Each stage is timed so you can see the computational cost next to the
ecological payoff. Stages 3 and 4 are also wrapped up in the one-call
`mt_clean_track()` entry-point, shown at the end.

## Meet the bird

**Pettstadt1 (AHH16, eObs tag 14053)** is a white stork nestling that
fledged from a nest near Pettstadt, Bavaria. The tag was deployed on
2024-06-13 by Wolfgang Fiedler (Max Planck Institute of Animal
Behavior). A representative subset of the track is bundled with
`move2utils` as `inst/extdata/Pettstadt1-14053.csv.gz`. You can
download the full track live with:

```r
library(move2)
movebank_store_credentials()  # run once, interactive
x <- movebank_download_study(
  study_id = 24442409,
  sensor_type_id = "gps",
  individual_local_identifier = "Pettstadt1 (AHH16, 14053)"
)
# x now carries the Movebank columns we need for quality filtering:
# gps_satellite_count, gps_dop, eobs_horizontal_accuracy_estimate
```

Bavarian white storks are partial migrants. We expect them in central
Europe in summer and autumn, with some wintering in Iberia, North
Africa, or the Sahel. Anything well outside that range is not the
stork — it is a fix error.

## Setup

```{r load, message = FALSE}
library(move2)
library(sf)
library(ggplot2)
library(rnaturalearth)
library(move2utils)

f <- system.file("extdata", "Pettstadt1-14053.csv.gz",
                  package = "move2utils")
raw <- mt_read(f)
cat(sprintf("raw: %d fixes, %d columns\n", nrow(raw), ncol(raw)))
```

## Stage 1 — Structural cleaning

Empty geometries (a GPS timestamp written without a position lock) and
duplicate timestamps (left behind when a Movebank download is resumed)
have to go before anything downstream can work. Nothing clever here —
just mechanical cleanup.

```{r basic-clean}
n0 <- nrow(raw)

x <- raw[!st_is_empty(raw), ]
n_empty <- n0 - nrow(x)

x <- mt_filter_unique(x, criterion = "first")
n_dup <- n0 - n_empty - nrow(x)

cat(sprintf("dropped %d empty + %d duplicate; kept %d\n",
             n_empty, n_dup, nrow(x)))
```

## Stage 2 — GPS-quality pre-filter

A GPS fix needs at least four satellites to exist at all, but fixes
right at that four-satellite edge have poor error geometry and often
land thousands of kilometres from the true position. The usual
telemetry rule of thumb is to keep `sat >= 5`. DOP and the
horizontal-accuracy estimate add independent evidence when those
columns are present.

```{r quality-filter}
xq <- mt_filter_gps_quality(x)
```

A quick before/after map shows how much these low-quality fixes matter.

```{r qmap, fig.width = 12, fig.height = 4.5, fig.cap = "Coordinate extent before and after mt_filter_gps_quality(). Left: raw fixes span three continents because of a handful of low-satellite teleports. Middle: same extent after filtering, with the dots collapsed onto Europe. Right: the same kept fixes zoomed to the European flyway, the first view in which the track's real shape is legible."}
world <- ne_countries(scale = "small", returnclass = "sf")

before_sf <- x
after_sf  <- xq

world_x <- c(-100, 80)
world_y <- c(-25, 80)
europe_x <- c(-15, 45)
europe_y <- c(30, 55)

p_before <- ggplot() +
  geom_sf(data = world, fill = "grey95", colour = "grey80",
          linewidth = 0.2) +
  geom_sf(data = before_sf, size = 1, alpha = 0.9,
          colour = "firebrick") +
  coord_sf(xlim = world_x, ylim = world_y) +
  labs(title = "Before filter", x = NULL, y = NULL) +
  theme_minimal(base_size = 10)

p_after <- ggplot() +
  geom_sf(data = world, fill = "grey95", colour = "grey80",
          linewidth = 0.2) +
  geom_sf(data = after_sf, size = 0.5, alpha = 0.8,
          colour = "steelblue4") +
  geom_rect(aes(xmin = europe_x[1], xmax = europe_x[2],
                ymin = europe_y[1], ymax = europe_y[2]),
            fill = NA, colour = "black", linewidth = 0.3) +
  coord_sf(xlim = world_x, ylim = world_y) +
  labs(title = "After sat>=5, DOP<=10, hacc<=100m",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 10)

p_after_zoom <- ggplot() +
  geom_sf(data = world, fill = "grey95", colour = "grey80",
          linewidth = 0.25) +
  geom_sf(data = after_sf, size =0.5, alpha = 0.5,
          colour = "steelblue4") +
  coord_sf(xlim = europe_x, ylim = europe_y) +
  labs(title = "After filter, zoomed to Europe",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 10)

if (requireNamespace("patchwork", quietly = TRUE)) {
  patchwork::wrap_plots(p_before, p_after, p_after_zoom, ncol = 3)
} else {
  print(p_before); print(p_after); print(p_after_zoom)
}
```

The extreme range collapses onto the expected European and
Mediterranean envelope; the zoomed panel resolves the structure that
the world-scale view squashes into dots, and it is the first view where
you can actually read the track. The filter does most of its work on
low-satellite fixes; DOP and the horizontal-accuracy estimate add
smaller gains.

## Stage 3 — Path-position check (bridge)

`mt_flag_outliers_bridge()` scores each fix by how far it sits from the
time-weighted average of its neighbours (a Brownian bridge between
them), scaled by the bridge width. The width depends only on the
timestamps, not on the positions, so a bad fix cannot widen its own
tolerance — it finds the geometric outliers that survive the quality
filter.

The default `method = "combined"` computes two scores in one pass — the
plain distance residual (dBBMM) and the across-track residual (dBGB) —
applies the threshold to each separately, and flags a fix if either
fires. On synthetic ground-truth tests this beats either score alone.

```{r bridge}
x_utm <- st_transform(xq, 32632)  # UTM 32N, central Europe

r_br <- mt_flag_outliers_bridge(x_utm,
                                 threshold_type = "entropy",
                                 iterations = 1,
                                 plot = FALSE)
cat(sprintf("bridge flagged %d of %d fixes\n",
             sum(r_br$is_outlier), nrow(r_br)))
```

The `bridge_eta`, `bridge_eta_para`, and `bridge_eta_perp` columns hold
the combined, along-track, and across-track scores for every fix —
handy for diagnosing the kind of error later.

```{r bridge-map, fig.cap = "Bridge-residual score in UTM coordinates. Flagged fixes (black circles) sit where a point is spatially inconsistent with its neighbours given the local sampling rate."}
r_br_ll <- st_transform(r_br, 4326)
r_br_ll$log_eta <- log10(pmax(r_br_ll$bridge_eta, 1e-10) + 1)

ggplot() +
  geom_sf(data = world, fill = "grey95", colour = "grey80",
          linewidth = 0.15) +
  geom_sf(data = r_br_ll, aes(colour = log_eta),
          size = 0.3, alpha = 0.6) +
  geom_sf(data = r_br_ll[r_br_ll$is_outlier, ],
          shape = 1, colour = "black", size = 2.2, stroke = 0.4) +
  scale_colour_viridis_c(option = "viridis",
                         name = expression(log[10](1 + eta))) +
  coord_sf(xlim = c(-15, 45), ylim = c(10, 60)) +
  labs(title = "mt_flag_outliers_bridge() — combined method, entropy",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 10)
```

## Stage 4 — Unusual-movement check

`mt_flag_outliers()` complements the path-position check by scoring
each fix against the joint distribution of step length, turning angle,
and their gap-aware auto-differences. It catches fixes whose movement
(speeds, turns, accelerations) does not match the animal's usual
behaviour, even when the point sits in a believable place.

It is good practice to drop the bridge-flagged fixes first: they would
otherwise distort the distribution the check builds from.

```{r probability}
xp <- r_br[!r_br$is_outlier, ]  # bridge-cleaned
xp <- st_transform(xp, 4326)    # this check ignores the CRS

r_prob <- mt_flag_outliers(xp, threshold_type = "entropy",
                            iterations = 1, plot = FALSE)
cat(sprintf("probability flagged %d of %d fixes\n",
             sum(r_prob$is_outlier), nrow(r_prob)))
```

## Cleaned track

The fixes flagged by the path-position and unusual-movement checks
together make up the removed set. Plotting the kept track next to the
removed fixes on a European extent shows what the pipeline caught.

```{r cleaned, fig.width = 9, fig.height = 5, fig.cap = "Left: fixes kept after the full pipeline. Right: fixes removed by stages 2–4, coloured by the stage that caught them."}
## merge removal sources
quality_removed <- x[!seq_len(nrow(x)) %in%
                      which(seq_len(nrow(x)) %in%
                             match(st_geometry(xq), st_geometry(x))), ]
## simpler: track in/out per stage via row-count accounting in the
## diagnostic only (exact mapping not needed for the map)

kept <- r_prob[!r_prob$is_outlier, ]

p_kept <- ggplot() +
  geom_sf(data = world, fill = "grey95", colour = "grey80",
          linewidth = 0.2) +
  geom_sf(data = kept, size = 0.2, alpha = 0.4,
          colour = "steelblue4") +
  coord_sf(xlim = c(-15, 40), ylim = c(25, 60)) +
  labs(title = sprintf("Cleaned track (n = %d)", nrow(kept)),
       x = NULL, y = NULL) +
  theme_minimal(base_size = 10)

removed_bridge <- r_br[r_br$is_outlier, ]
removed_prob   <- r_prob[r_prob$is_outlier, ]

p_removed <- ggplot() +
  geom_sf(data = world, fill = "grey95", colour = "grey80",
          linewidth = 0.2) +
  geom_sf(data = st_transform(removed_bridge, 4326),
          size = 1.2, colour = "firebrick") +
  geom_sf(data = removed_prob,
          size = 1.2, colour = "orange") +
  coord_sf(xlim = c(-15, 40), ylim = c(25, 60)) +
  labs(title = sprintf("Removed by bridge (red, n=%d) and probability (orange, n=%d)",
                         nrow(removed_bridge), nrow(removed_prob)),
       x = NULL, y = NULL) +
  theme_minimal(base_size = 10)

if (requireNamespace("patchwork", quietly = TRUE)) {
  patchwork::wrap_plots(p_kept, p_removed, ncol = 2)
} else {
  print(p_kept); print(p_removed)
}
```

## The one-call alternative — `mt_clean_track()`

The pipeline above narrates each check so you can see what is happening
and why. For routine cleaning, `mt_clean_track()` runs all four checks
behind a single call: it applies the path-position, there-and-back
(detour), unusual-movement, and speed checks, combines them with the
default consensus rule, removes groups of bad fixes that form isolated
blocks, and iterates until nothing new is flagged.

The default rule is `consensus = "evidence_corroborated"`: it weighs
each check's evidence on a common scale and flags a fix when the
combined evidence is positive **and** either at least two checks agree
or the there-and-back check alone is overwhelmingly sure. You can pick
a different rule via `consensus =` — the earlier behaviour is
`"class_aware"`, and `"strict"`, `"majority"`, `"speed_trusted"`,
`"weighted_evidence"`, `"any"`, and `"custom"` are also available (see
`?mt_flag_consensus`). Whichever rule you choose, the `error_class`
column — naming which kind of error caused each flag — is always
computed.

White storks have a known sustained-flight ceiling around 50 m/s.
Supplying that as a physiological cap lets the speed check peel off any
spoof- or teleport-class clusters cleanly before the other checks run
on the survivors.

```{r clean-track}
clean_one_call <- mt_clean_track(xq, v_max = 50, plot = FALSE)
cat(sprintf("mt_clean_track kept %d of %d fixes\n",
             nrow(clean_one_call), nrow(xq)))
```

The kept set should be close to the union of the bridge and probability
flags from the staged pipeline; the consensus rule and block expansion
give it slightly different behaviour on edge cases. Use whichever style
suits you — the staged version when you want to inspect each signal,
the one-call version when you trust the defaults and want to move on.

## Composition summary

```{r perf-summary}
tab <- data.frame(
  stage = c("1. structural cleaning",
            "2. GPS-quality filter",
            "3. bridge (combined)",
            "4. probability (entropy)"),
  n_in = c(n0, nrow(x), nrow(xq), nrow(xp)),
  n_out = c(nrow(x), nrow(xq), sum(!r_br$is_outlier),
             sum(!r_prob$is_outlier))
)
tab$n_dropped <- tab$n_in - tab$n_out
print(tab, row.names = FALSE)
```

Each stage handles a distinct kind of error, and each is cheap:

- Structural cleaning is O(n) and mechanical.
- The GPS-quality filter is O(n) and kills the most pathological fixes
  at source — often the difference between a usable and an unusable
  track.
- The bridge check is O(n) with a single neighbour scan.
- The unusual-movement check is O(n) with a fully vectorised density
  lookup; a cap on the 2D turn/step histogram bin count keeps the
  `terra` raster work linear in n even on heavy-tailed step
  distributions.

## Notes on reproducibility

- The bundled CSV was made by filtering the Movebank download to a
  stable subset of columns. All three GPS-quality columns
  (`gps-satellite-count`, `gps-dop`,
  `eobs-horizontal-accuracy-estimate`) are kept.
- You can re-download the full Movebank study with
  `move2::movebank_download_study(24442409, individual_local_identifier = "Pettstadt1 (AHH16, 14053)")`
  after running `move2::movebank_store_credentials()` once.
- Study acknowledgement: *LifeTrack White Stork Bavaria*, data
  contributed by Wolfgang Fiedler and colleagues, Max Planck Institute
  of Animal Behavior.

## Further reading

- `vignette("OUTLIER_1_getting_started", package = "move2utils")` — the
  unified `mt_clean_track()` workflow and a brief tour of all four
  checks.
- `vignette("OUTLIER_2_diagnose_clean_track", package = "move2utils")` — the
  post-run health check.
- `vignette("OUTLIER_3_state_conditional", package = "move2utils")` — when the
  diagnostic flags two behaviours, the recipe for cleaning each
  behavioural state separately.
- `vignette("OUTLIER_4_outlier_bridge", package = "move2utils")` — the
  path-position check and the directional error-morphology classifier;
  for users who want fine control over just one check.
- `vignette("OUTLIER_5_persistence_score", package = "move2utils")` —
  multi-scale annotation that scores how confident each flag is; useful
  as a post-cleaning confidence filter.
- `vignette("OUTLIER_heterogeneous_error_regimes", package = "move2utils")` — outlier detection with mixed error regimes, one sensor at a time.
- `vignette("OUTLIER_example_leo_migration", package = "move2utils")` — outlier
  detection on irregular, large-scale satellite data.
