---
title: "Heterogeneous error regimes: one sensor at a time"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Heterogeneous error regimes: one sensor at a time}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 9,
  fig.height = 5,
  message  = TRUE
)
```

## What these checks assume

All the cleaning in this package rests on one assumption: each fix is a
noisy estimate of where the tracked device — and so the animal carrying
it — actually was at that moment. Every check inside `mt_clean_track()`
— path position (`mt_flag_outliers_bridge()`), there-and-back spikes
(`mt_flag_outliers_detour()`), unusual movement (`mt_flag_outliers()`),
and speed (`mt_flag_speed_cap()`) — reads neighbouring fixes as samples
from one underlying path and asks which ones do not belong on it. This
is the *Lagrangian* view: we follow the moving body.

Some sensor streams report detections at fixed receivers rather than
positions of the animal — Sigfox-geolocation, LoRa-network
triangulation, VHF antenna scans, camera-trap captures, hydrophone
events. Their raw output describes the receiving infrastructure, not the
animal, so it breaks the assumption. This is the *Eulerian* view.
Exclude these streams from cleaning; they belong in occupancy,
detection-radius, or hidden-Markov models downstream.

What matters is the *representation*, not the original sensor. If an
Eulerian network has already been turned into a path — a trajectory
reconstructed from a camera array with individual re-identification, or a
chronological sequence of geolocated photographs of a known individual —
that reconstructed path is exactly what these checks are built to clean.

The rest of this vignette assumes Lagrangian data and covers one harder
case: a single track that interleaves more than one Lagrangian sensor at
very different precisions.

## When you need this

Modern deployments increasingly mix Lagrangian sensors at very different
precisions within one track. A GPS + GLS deployment gives metre-scale
fixes when the GPS gets a satellite lock and hundred-kilometre fixes from
twilight estimates when it does not. An Argos-only deployment mixes
`LC = 3` fixes (a few hundred metres of error) with `LC = B` fixes (no
useful accuracy). GPS + Argos Doppler deployments span three orders of
magnitude in nominal positional error from one fix to the next.

Each check fits its threshold from the spread of the events it is handed.
On a mixed track that spread has two (or more) humps, and any single
threshold either lets the kilometre-scale events through (because it
respects the broad tail) or rejects the metre-scale tail (because it sits
inside the broad bulk). Neither is the answer you want.

The fix is not another knob on `mt_clean_track()`. It is to **split the
data by sensor before cleaning, clean each piece on its own terms, and
re-merge if your analysis needs the combined path.** The rest of this
vignette walks through that pattern.

## How to recognise the case

If the tag reports a `sensor_type_id` event column with more than one
value, you have a mixed-regime track. The clearest sign is two very
different orders of magnitude in nominal horizontal accuracy:

```{r demo-detect, eval = FALSE}
library(move2)
library(dplyr)

# `your_track` is the Movebank object you are inspecting.
table(your_track$sensor_type_id)
#>   <sensor_A>   <sensor_B>
#>        69860        72949
```

Look up each sensor code in the Movebank attribute dictionary or your
study metadata so you know what is what. Two Lagrangian sensors at very
different precisions — GPS plus GLS, or GPS plus Argos Doppler — are the
case this vignette addresses. A track that mixes a Lagrangian sensor with
an Eulerian one (Sigfox-geolocation, LoRa, antenna detection) is **not**
this case; exclude the Eulerian stream entirely (see the first section)
before going on.

Other signs of a mixed-regime track:

- `argos_lc` taking values across `3, 2, 1, 0, A, B, Z` in the same
  track.
- `eobs_horizontal_accuracy_estimate` (or any per-fix accuracy column)
  with a multi-humped distribution.
- The same animal carrying two devices that report into the same
  Movebank study with very different sampling cadence.

## Why one call to `mt_clean_track()` does the wrong thing

Hand a mixed-Lagrangian object straight to `mt_clean_track()` and every
threshold — path position, unusual movement, there-and-back, and speed —
sees a step-speed spread that mixes metre-scale steps from the precise
stream with kilometre- or hundred-kilometre jumps from the coarse one.
Whatever cap the auto-fitter picks sits either above the coarse bulk (so
almost no coarse fix is flagged, even an obviously wrong one) or between
the two bulks (so most coarse fixes are flagged, including the merely
imprecise ones).

Neither is a useful answer. `mt_diagnose_clean_track()` will usually show
the multi-humped step-speed distribution plainly, and that is your cue to
stop and pre-split.

## The pre-split / re-merge pattern

The split is mechanical: filter the move2 object on the per-fix sensor
column, clean each subset with its own `mt_clean_track()` call, then join
the per-event flags back onto the original object by `event_id`. Each
subset is internally homogeneous, so each check's assumptions hold within
the call.

```{r split, eval = FALSE}
library(move2utils)
library(move2)
library(dplyr)

# 1. Split by sensor (event-level column).
your_track_A <- your_track[your_track$sensor_type_id == "<sensor_A>", ]
your_track_B <- your_track[your_track$sensor_type_id == "<sensor_B>", ]

# 2. Clean each on its own.  Each call sees one error regime; the
#    threshold checks fit a single, well-behaved distribution.
#    See "Choosing the speed check" below for whether to pass
#    mass+mode (or a hard v_max) to each call.
clean_A <- mt_clean_track(your_track_A, plot = FALSE, remove = FALSE,
                          silent = TRUE)
clean_B <- mt_clean_track(your_track_B, plot = FALSE, remove = FALSE,
                          silent = TRUE)

# 3. Re-merge by joining the per-event flag back onto the original
#    object using `event_id`.  The two sub-objects may share track
#    IDs (e.g. when a single deployment had both sensors), so a
#    naive rbind() of the move2 objects can error on duplicates.
#    An event_id join is the safe pattern.
flags <- rbind(
  data.frame(event_id = clean_A$event_id, is_outlier = clean_A$is_outlier),
  data.frame(event_id = clean_B$event_id, is_outlier = clean_B$is_outlier)
)
cleaned <- your_track
cleaned$is_outlier <- flags$is_outlier[match(cleaned$event_id, flags$event_id)]
```

Points worth noting:

- **Split on an event-level column.** Track-level columns (everything in
  `mt_track_data()`) describe the whole deployment and cannot resolve a
  within-track sensor mix. Use a column that varies per fix
  (`sensor_type_id`, `argos_lc`, or one you derived from per-fix
  accuracy).
- **Each subset is a valid move2.** Subsetting with `[` keeps the move2
  class, the time column, and the track-id column.
- **Re-merge with an `event_id` join, not `rbind`.** When a single
  deployment carried both sensors, the two subsets share track IDs and
  `rbind()` refuses to combine them. Joining the flag back onto the
  original `your_track` by `event_id` sidesteps this, and also carries
  the flag onto events from any tracks you filtered out upstream (which
  then hold `NA` for `is_outlier`).

## Choosing the speed check for each sub-stream

After the split, each sub-stream gets its own `mt_clean_track()` call.
The biggest per-call decision is whether to give it a physiological speed
cap — `mass + mode` for the allometric cap from `v_phys_estimate()`, or a
hard `v_max =` in m/s — or to let the speed check's relative auto-cap
work alone. The rule is mechanical once you state the comparison:

> Compare the sensor's noise floor in step-speed units
> (`positional_error / sampling_interval`) against the animal's
> physiological top step speed. If the noise floor is *well below*
> physiology, supply a physiological cap. If it is *at or above*
> physiology, do not.

Two worked cases:

- **GPS, 10 m typical error, 10-minute sampling.** Noise floor
  ≈ 10 m / 600 s ≈ 0.017 m/s, far below any terrestrial-mammal
  maximum. A physiological cap (say 14 m/s for a ~700 kg running animal)
  cleanly separates impossible steps from real movement, and the
  block-detection step's absolute cap catches outliers the relative
  auto-cap — fitted from the same distribution that holds the outliers —
  might miss.
- **GLS twilight, ~100 km typical error, daily fixes.** Noise floor
  ≈ 100 km / 86400 s ≈ 1.2 m/s, already at or above the sustained speed
  of most species the sensor is fitted to. A physiological cap here would
  flag the bulk of the GLS data as "impossible" when those step speeds
  are just normal sensor noise over slow daily displacement. The relative
  auto-cap is the right tool: it fits the threshold from the sensor's own
  step speeds, so it lands at the right scale for the actual noise.

The same reasoning covers other sensor pairs:

- **Argos LC 3** (~250 m) on a soaring raptor: noise floor is tiny next
  to the raptor's sustained ground speed; a physiology cap is fine.
- **Argos LC B** (km-scale, sometimes unbounded) on the same raptor: a
  physiology cap over-flags; rely on the auto-cap.
- **Argos Doppler shift** (~1 km typical error, hourly cadence):
  borderline. Run `mt_diagnose_clean_track()` and look at the step-speed
  distribution. If the bulk sits below your expected physiological max,
  use the physiology. If it straddles or sits above, use the auto-cap.

When in doubt, run both and compare. A physiology cap should only remove
fixes above a defensible biological maximum; if it flags a large fraction
of fixes, the cap is wrong for that sub-stream (or its error regime is
more variable than you thought, in which case the auto-cap is the safe
choice).

## What if I have multi-individual mixed-sensor data?

This is where nested `pool_by` comes in. Suppose 50 raptors are tagged,
each tag reports both GPS and Argos Doppler fixes, and you want
population-strength thresholds while keeping the flag union scoped to each
individual. Split by sensor first, then clean each homogeneous subset
with a two-element `pool_by`:

```{r nested, eval = FALSE}
# `your_track` is the multi-individual, mixed-sensor object you
# downloaded from Movebank. Split it by sensor first:
your_track_gps   <- your_track[your_track$sensor_type_id == "<gps_code>", ]
your_track_argos <- your_track[your_track$sensor_type_id == "<argos_doppler_code>", ]

# Each homogeneous sub-object is then cleaned with nested pool_by.
# "study_id" supplies the fit distribution (all raptors contribute);
# "individual_local_identifier" is the union scope (a flag added by
# the pool is scoped to that animal's tracks only).
clean_gps <- mt_clean_track(
  your_track_gps,
  pool_by = c("study_id", "individual_local_identifier"),
  plot = FALSE, remove = FALSE
)
clean_argos <- mt_clean_track(
  your_track_argos,
  pool_by = c("study_id", "individual_local_identifier"),
  plot = FALSE, remove = FALSE
)
```

The two-element form has two roles. The outer column (`"study_id"`) says
where the threshold-fitting distribution comes from — all raptors
contribute. The inner column (`"individual_local_identifier"`) says how
far a pooled flag reaches — only that animal's tracks. Two roles, two
columns. (See `?mt_clean_track` for the full semantics.)

A deeper hierarchy — species, population, individual, tag — is not
supported. `pool_by` has exactly two roles; with more nesting levels you
pick the pair that captures the claim you want to make (whose
distribution do I trust; within which level should the flag union act). A
third level would only pay off under hierarchical / partial-pooling
threshold estimation, which the cascade does not do.

## Edge cases

**Time gaps introduced by splitting.** Filtering by sensor naturally
opens longer time gaps in each subset (only that sensor's rows remain).
The path-position check handles this — it is gap-aware by construction.
So does the unusual-movement check, which normalises by time lag. The
there-and-back check is time-insensitive and unaffected. The speed check
sees the subset's own step speeds, which are still physically meaningful
(steps within one sensor stream).

**Sensors that share an error regime.** Two sensors with the same error
regime (say two GPS manufacturers, each ~10 m accuracy) need no split:
the mix is homogeneous in the only sense that matters here. Split when the
underlying distribution is bimodal, not just when the column has two
values.

**A sensor whose error is encoded per-fix (e.g. Argos LC).** You can
treat each LC class as its own stream and split four ways. More often,
you supply the per-fix sigma directly through the path-position check's
`location_error =` argument; it then inflates its tolerance for
high-error fixes and you keep all classes in one call. Split by sensor
when the error is *categorical* (which device produced this fix); use
`location_error =` when the error is *quantitative* (this fix has this
sigma).

## Summary

- `mt_clean_track()` assumes one error regime per call, and assumes that
  error regime is Lagrangian.
- Mixed-Lagrangian tracks should be **pre-split by sensor on an
  event-level column**, cleaned independently, and re-merged via an
  **`event_id` join** onto the original object.
- For multi-individual datasets, nested `pool_by` after the split gives
  cohort-strength threshold fitting with an individual-scoped flag union.
- The two-level cap on `pool_by` is deliberate: it has two semantic
  roles, not many.

## Further reading

- `vignette("OUTLIER_1_getting_started", package = "move2utils")` — the
  unified `mt_clean_track()` workflow and a brief tour of all four
  checks.
- `vignette("OUTLIER_2_diagnose_clean_track", package = "move2utils")` — the
  post-run health check.
- `vignette("OUTLIER_3_state_conditional", package = "move2utils")` — when the
  diagnostic flags bimodal behaviour, the recipe for cleaning each
  behavioural state separately.
- `vignette("OUTLIER_4_outlier_bridge", package = "move2utils")` — the
  path-position check and the directional error-morphology classifier;
  for users who want fine-grained control over just one check.
- `vignette("OUTLIER_5_persistence_score", package = "move2utils")` —
  multi-scale annotation that scores how confidently each flag is an
  outlier; useful as a post-cleaning confidence filter.
- `vignette("OUTLIER_example_outlier_whitestork", package = "move2utils")` — a full
  narrated cleaning pipeline on a real high-frequency stork track.
- `vignette("OUTLIER_example_leo_migration", package = "move2utils")` — outlier
  detection on irregular, large-scale satellite data.
