---
title: "Getting started: cleaning a track with move2utils"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started: cleaning a track with move2utils}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 4,
  message = TRUE
)
```

## Why clean a track?

Almost every GPS or satellite track contains a few bad locations: a fix
from the wrong place, a brief satellite-geometry error, a tag switched
on before it was fitted. They are usually a tiny fraction of the data,
but they do real damage — one location far out of place inflates every
speed you compute and distorts home ranges. Cleaning is the first step
of most movement analyses.

There is no definitive list of correct outliers. An outlier is a
judgement that a location is too implausible to keep; some are obvious,
others borderline. So `move2utils` does not try to find a perfect set.
It flags the *conspicuous* errors reliably, leaves believable data
alone, and shows you what it did so the judgement stays yours.

> Always plot the result and check it. No automatic cleaner should be
> trusted blind.

`mt_clean_track()` does the cleaning by combining four independent
checks (path position, there-and-back spikes, unusual movement, and
speed) into a single decision. This vignette covers the one-call
workflow, then each check on its own for finer control.

## Load data

We use the `CPF_A` track from the package's bundled synthetic data — a
1748-fix central-place-forager simulation with bad locations injected at
known positions, so we can see what "clean" should look like. The same
track is the demonstration fixture in vignettes 3, 4 and 5. For narrated
cleaning of real GPS tracks, see the `OUTLIER_example_*` vignettes under
"Where to go next".

```{r load-data, message = FALSE}
library(move2)
library(sf)
library(move2utils)

path   <- system.file("extdata/synthetic_tracks.csv.gz", package = "move2utils")
tracks <- mt_read(path)
cpf_a  <- filter_track_data(tracks, .track_id = "CPF_A")
cat(nrow(cpf_a), "locations\n")
```

If you pull your data from Movebank, see the appendix for the
recommended download pattern — it matters which attribute columns you
ask for, because the cleaning can use Movebank's per-fix quality
information to sharpen its decisions.

## The one-call workflow

`mt_clean_track()` runs the four checks in turn, decides which locations
to flag, removes groups of bad locations that form isolated blocks, and
iterates until nothing new is flagged. By default it returns the cleaned
track.

```{r clean-default, fig.width = 10, fig.height = 5}
clean <- mt_clean_track(cpf_a)

cat("kept", nrow(clean), "of", nrow(cpf_a),
    "  (removed", nrow(cpf_a) - nrow(clean), ")\n")
```

The messages are the function narrating its decisions; the figure shows
the kept track with the removed locations marked. Check that plot: do
the removed points match where you would have pointed?

**How the decision is made.** Each of the four checks occasionally fires
on good data (a real sharp turn, a genuine fast burst), so the cleaner
does not simply count votes. It weighs how strongly each check distrusts
a location and flags it when the combined evidence is positive **and**
either two checks agree or one is overwhelmingly sure. This catches
clear errors that a single check happens to see, without over-cleaning
believable data. This rule is the default (`consensus =
"evidence_corroborated"`); you can change it via `consensus =` — the
earlier behaviour is `"class_aware"`, and `"strict"`, `"majority"`,
`"speed_trusted"`, `"weighted_evidence"`, `"any"` and `"custom"` are
also available (see `?mt_flag_consensus`).

To see *which* check caught *what*, set `remove = FALSE`: every location
is kept and the ones that would be removed are labelled.

```{r clean-inspect}
flagged <- mt_clean_track(cpf_a, remove = FALSE, plot = FALSE)

table(flagged$is_outlier)

table(path    = flagged$flagged_by_bridge,
      back    = flagged$flagged_by_detour,
      unusual = flagged$flagged_by_prob,
      fast    = flagged$flagged_by_speed)
```

The columns `flagged_by_bridge` (path position), `flagged_by_detour`
(there-and-back), `flagged_by_prob` (unusual movement), and
`flagged_by_speed` (speed), together with `flag_iteration` and
`block_id`, record the provenance of each flag. The `error_class` column
names which rule caused the flag (e.g. a clear there-and-back spike, an
impossible speed, or membership of a removed block) — see
`vignette("OUTLIER_2_diagnose_clean_track", package = "move2utils")` for
the taxonomy.

Errors also come in **groups**: a run of fake locations from a spoof, or
a cluster stuck at one wrong place. The cleaner detects locations
forming a block cut off from the real trajectory and removes the whole
block, not just its edges.

## When to supply a physiological speed cap

By default `v_max = NULL`, and the speed check infers a plausible cap
from the track itself. This is safe on most species, but it has a known
limit: long, internally consistent error clusters (typical of GPS
spoofs or multi-hour position jumps) have small internal step speeds and
only stand out at their edges. A single fixed physiological cap, applied
repeatedly until nothing more is removed, dissolves them cleanly.

The block-detection step also uses `v_max` to decide which locations are
connected. It only removes a block when the data show the expected
"trajectory plus isolated clusters" shape; if the inferred cap sits
inside the species' normal speed range, removing blocks would cut real
trajectory, so the step is declined automatically (the decision is
printed in the narration). To enable it on a fast-flying species, give a
physiological cap via `v_max =` or the `(mass, mode)` route below.

```{r v-max, eval = FALSE}
## a large eagle sustains roughly 30 m/s, a stork roughly 50 m/s
clean <- mt_clean_track(track, v_max = 30)
```

With `v_max` supplied, `mt_clean_track()` first peels locations above
that cap to convergence (`mt_peel_speed()`), then runs the remaining
checks on what survives. `mt_suggest_speed_cap()` is a diagnostic
helper: it reports the data-driven candidates and warns when they
disagree, but deliberately does not pick `v_max` for you.

```{r suggest-cap, eval = FALSE}
mt_suggest_speed_cap(track)
```

### A principled default from body mass and locomotor mode

Without a published speed cap, you can derive one from body mass and
locomotor mode using the Hirt et al. (2017) scaling law.
`v_phys_estimate(mass, mode)` returns the predicted maximum sustained
speed in m/s with a confidence interval; `mt_clean_track()` accepts the
same `(mass, mode)` pair directly.

```{r v-phys, eval = FALSE}
v_phys_estimate(mass = 5, mode = "flying")            # eagle-class, printed with CI
clean <- mt_clean_track(track, mass = 5, mode = "flying")
```

`mt_suggest_speed_cap()` takes the same arguments and overlays the
allometric prediction on the step-speed distribution alongside any
user value, so the empirical break, the allometric prediction, and a
supplied `v_max` appear on one plot. Disagreement is itself
informative: an empirical break well above the allometric prediction
signals contamination; one well below it is within-distribution
structure (behavioural states, gait changes) that should not be cut.

```{r suggest-triangulated, eval = FALSE}
mt_suggest_speed_cap(track, mass = 5, mode = "flying", v_max = 30)
```

If the locomotor mode is uncertain (a stork that flies and walks, a seal
that swims and walks), pass `mass` alone and the diagnostic shows all
three mode-specific lines.

```{r suggest-three-modes, eval = FALSE}
mt_suggest_speed_cap(track, mass = 1)
```

### Where to find body mass

Body mass is the only species-level input the allometric helper needs.

- **Movebank reference data** — per-individual mass is often in the
  metadata:
  ```r
  ref <- move2::mt_track_data(track)
  ref$animal_mass   # kg, when populated
  ```
- **Trait databases** — EltonTraits 1.0
  ([Wilman et al. 2014](https://doi.org/10.1890/13-1917.1)) for birds
  and mammals; PanTHERIA
  ([Jones et al. 2009](https://doi.org/10.1890/08-1494.1)) for mammals;
  Amniote ([Myhrvold et al. 2015](https://doi.org/10.1890/15-0846R.1))
  for birds, mammals and reptiles. Download the CSV once, look up by
  Latin name.
- **`traitdataform`**
  ([repo](https://github.com/EcologicalTraitData/traitdataform))
  aggregates these under one schema.
- **Wikidata SPARQL** — query `wdt:P2067` (body mass) for a species'
  Q-id; lightest live route, coverage varies.

## The individual checks

Each check is also exported on its own, for inspecting a single signal
or building your own pipeline.

**Path position — `mt_flag_outliers_bridge()`.** Scores each location by
how far it sits from where its temporal neighbours place it, scaled by a
width that depends only on the timestamps — so a bad location cannot
inflate its own tolerance. A directional variant separates along-track
from across-track deviation and classifies the kind of error. See
`vignette("OUTLIER_4_outlier_bridge", package = "move2utils")`.

```{r bridge-only, fig.width = 10, fig.height = 5}
br <- mt_flag_outliers_bridge(cpf_a, plot = FALSE)
cat("path-position flagged:", sum(br$is_outlier), "\n")
```

**Unusual movement — `mt_flag_outliers()`.** Scores each location by how
unusual its step length and turn are relative to the animal's own
movement, catching kinematically implausible fixes the geometry alone
would miss. `threshold_type = "gap"` (default) finds a break in the
log-probability tail; `"entropy"` is more conservative;
`"significance"` and `"percentile"` are the older z-score and quantile
rules.

```{r prob-only, fig.width = 10, fig.height = 5}
pr <- mt_flag_outliers(cpf_a, plot = FALSE)
cat("unusual-movement flagged:", sum(pr$is_outlier), "\n")
```

**Speed — `mt_flag_speed_cap()`.** Flags steps whose implied speed
exceeds a threshold. The threshold is data-driven by default (`auto`);
pass `threshold_type = "hard"` with `v_max` for a fixed cap, or
`mt_peel_speed(x, v_max)` for the iterative peel that dissolves coherent
clusters.

```{r speed-only, fig.width = 10, fig.height = 5}
sc <- mt_flag_speed_cap(cpf_a, plot = FALSE)
cat("speed flagged:", sum(sc$is_outlier), "\n")
```

The there-and-back check is `mt_flag_outliers_detour()`; the rule that
combines the four into one decision is `mt_flag_consensus()`.

## Persistence and other variants

`mt_persistence_score()` annotates any flagged output with a confidence
score: how anomalous each flagged location still looks when viewed over
wider time windows (`scales = c(2, 4, 8)` by default). It does not
change `is_outlier`; it adds a `persistence_count` column from 1
(flagged only at native resolution) to 4 (flagged at every scale). It is
useful as a confidence filter on cascade output — see
`vignette("OUTLIER_5_persistence_score", package = "move2utils")`.

```{r persistence}
clean     <- mt_clean_track(cpf_a, plot = FALSE, remove = FALSE)
annotated <- mt_persistence_score(clean, silent = TRUE)
print(table(annotated$persistence_count[annotated$is_outlier]))
```

`mt_sequential_outliers()` and `mt_combined_outliers()` provide
sequential-scan and majority-vote variants of the movement check; see
their help pages.

## A health check

`mt_diagnose_clean_track()` adds a panel of views that catch what a
single map hides — an animal with two behaviours (resting versus
commuting) that may need cleaning separately, or a run that never
settled. Panels that look wrong print a short interpretive note.

```{r diagnose, fig.width = 11, fig.height = 7, eval = FALSE}
res <- mt_clean_track(track, mass = 3.4, mode = "flying",
                      plot = FALSE, remove = FALSE)
mt_diagnose_clean_track(res)
```

A full walkthrough on real data is in
`vignette("OUTLIER_2_diagnose_clean_track", package = "move2utils")`.

## A sensible workflow

1. Load the track; drop empty geometries and duplicate timestamps.
2. If the tag reports fix quality, run `mt_filter_gps_quality()` (see
   the appendix).
3. Run `mt_clean_track()`.
4. Plot the result. If you know the animal's top speed, re-run with
   `v_max =` (or `mass =`, `mode =`).
5. Refine with the per-check functions only if the unified call misses
   something specific.
6. Run `mt_diagnose_clean_track()` to confirm the run was healthy.

## Where to go next

Read in order: this vignette, then `OUTLIER_2` to interpret your run,
then `OUTLIER_3` if your animal has multiple behaviours. The rest go
deeper as needed.

- `vignette("OUTLIER_2_diagnose_clean_track", package = "move2utils")` —
  the after-cleaning health check.
- `vignette("OUTLIER_3_state_conditional", package = "move2utils")` —
  cleaning different behaviours separately.
- `vignette("OUTLIER_4_outlier_bridge", package = "move2utils")` — the
  path-position check and the kinds of error it distinguishes.
- `vignette("OUTLIER_5_persistence_score", package = "move2utils")` —
  scoring how confident each flag is.
- `vignette("OUTLIER_6_troubleshooting", package = "move2utils")` — what
  to do when the default cleaning removes too much, misses errors, or
  errors out.
- `vignette("OUTLIER_heterogeneous_error_regimes", package = "move2utils")` —
  tracks mixing sensors with different error, cleaned one sensor at a
  time.
- `vignette("OUTLIER_example_outlier_whitestork", package = "move2utils")` —
  a full cleaning on a real high-frequency stork track.
- `vignette("OUTLIER_example_leo_migration", package = "move2utils")` —
  cleaning irregular, large-scale satellite data.

## Appendix: downloading your data from Movebank

The cleaning can use Movebank's per-fix quality information — how many
satellites a fix used, how precise it was, what kind of fix it was — to
sharpen decisions on borderline locations. You just need to ask for
those columns, as they are not all returned by default.

The principle: *if Movebank has the quality information, ask for it.*
Each quality column independently sharpens the decision on doubtful
fixes. If a column is missing the cleaning still runs —
`mt_filter_gps_quality()` skips criteria whose columns it can't find,
and the path-position check falls back to a supplied `location_error` —
but quality-informed flags are more reliable. The extra columns cost
little. Two things to do at download:

1. **Restrict to GPS** with `sensor_type_id = "gps"`. Many tags emit
   accelerometer bursts as separate rows with empty geometries; without
   this, a download can be several times larger and mostly empty rows.
2. **Name the quality columns** in `attributes`, so the data dependency
   is explicit and robust to studies with customised defaults.

Columns recognised out of the box:

| Movebank column | What it represents |
|---|---|
| `gps_satellite_count` / `gnss_satellite_count` | satellites used in the fix |
| `gps_hdop` / `gps_pdop` / `gps_dop` (or `gnss_*`) | dilution of precision |
| `gps_fix_type` / `gnss_fix_type` | 2D vs 3D fix |
| `eobs_horizontal_accuracy_estimate` | per-fix 1-σ accuracy in metres (e-obs) |
| `argos_lc` | Argos location class |

```{r movebank-pattern, eval = FALSE}
## everything the study has
track <- move2::movebank_download_study(
  study_id, sensor_type_id = "gps", attributes = "all")

## or name the columns explicitly
track <- move2::movebank_download_study(
  study_id, sensor_type_id = "gps",
  attributes = c("timestamp", "location_lat", "location_long",
                 "individual_local_identifier",
                 "gps_satellite_count", "gps_hdop", "gps_pdop", "gps_dop",
                 "gps_fix_type", "eobs_horizontal_accuracy_estimate",
                 "argos_lc"))
```

For modern multi-constellation tags the columns may start with `gnss_`
rather than `gps_`; where both are present, GNSS is preferred. For a
quick look without quality information, `attributes = NULL` returns the
minimum (timestamp, location, track id) and the pipeline runs in
degraded mode.

**First download from a study.** Movebank requires you to accept the
study's licence once. The first call fails with the licence terms and a
line like:

```
'license-md5'='306ac0a2292eb02b9b42d1b5faeca786'
```

Copy that string back as a named argument (quoted, because of the
hyphen):

```r
track <- movebank_download_study(
  study_id, sensor_type_id = "gps", attributes = c(...),
  "license-md5" = "306ac0a2292eb02b9b42d1b5faeca786")
```

You only do this once per study. If the object arrives with empty rows,
drop them before cleaning: `x <- x[!sf::st_is_empty(x), ]`.

**Non-Movebank vendors.** If your columns differ from the Movebank names
(Vectronic, Lotek, Telonics, Sirtrack, custom feeds), auto-detection
won't find them and the cleaning runs in degraded mode (still works,
without the extra information). A unified quality-column resolver is in
design (`DESIGN_quality_columns.md`); for now, either rename the columns
to the Movebank names before cleaning, or pass per-fix quality functions
directly via `mt_flag_outliers(quality_columns = list(...))`.
