---
title: "State-conditional outlier cleaning"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{State-conditional outlier cleaning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 9,
  fig.height = 5,
  message  = TRUE
)
```

## When you need this

`mt_clean_track()` runs four checks — path position
(`mt_flag_outliers_bridge()`), there-and-back spikes
(`mt_flag_outliers_detour()`), unusual movement
(`mt_flag_outliers()`), and speed (`mt_flag_speed_cap()`). Each one
compares every location against a single distribution: of bridge
residuals, detour ratios, movement probabilities, or step speeds.

That works when the animal moves in one consistent way. It breaks
down when the animal has several behaviours with very different
kinematics — resting at a perch, soaring, active flight, migration
glides. Each behaviour adds its own peak to the distribution, and a
single threshold then has to choose between two bad options:

- cut between the peaks, which over-flags the smaller behaviour, or
- cut beyond the largest peak, which misses the smaller behaviour's
  own outliers.

Neither is right. The fix is to *split the track by behaviour and
clean each piece on its own*, so every check sees a single peak and
can find the outlier tail relative to that one behaviour.

## How to recognise the case

`mt_diagnose_clean_track()` prints a clear note when this pattern is
present:

> Panel 1: 3 substantive modes detected at 0.01, 0.14, 6.91 m/s --
> bimodal behaviour. The per-fix detectors threshold against a single
> distribution; consider state-conditional analysis or filtering to
> one mode before cleaning.

Together with a flag rate that stays high over a contiguous stretch
of time (Panel 2), this is the most common problem users hit on
multi-behaviour species.

## The `state =` parameter

`mt_clean_track()` takes a `state =` argument. It splits each track
into runs of constant state and runs the full pipeline separately on
each run. The package treats your state labels as authoritative: how
you decide the states (speed threshold, HMM, BCPA, manual
annotation) is up to you, not the package.

There are three accepted forms.

1. **A column name on the move2 object**:

   ```{r api-column, eval = FALSE}
   x_clean <- mt_clean_track(x, v_max = 50, state = "behaviour")
   ```

2. **A per-fix vector** of length `nrow(x)`:

   ```{r api-vector, eval = FALSE}
   x_clean <- mt_clean_track(x, v_max = 50, state = my_state_vec)
   ```

3. **`NULL`** (default): one global distribution, the standard
   behaviour.

`NA` values in the state column or vector form their own state —
runs of `NA` are cleaned in their own context. Runs shorter than 3
fixes pass through unflagged, because the checks need at least 3
points to compute a residual or auto-difference.

## Where to get the state assignment from

### From a column already on the move2

The simplest case. If your data already carries a behavioural label
(annotated by the field team, derived from accelerometer, imported
from Movebank's annotation columns), pass the column name:

```{r from-column, eval = FALSE}
x_clean <- mt_clean_track(x, v_max = 50, state = "behaviour")
```

### From a speed threshold

The cheapest do-it-yourself segmentation. Pick a valley between
speed peaks (`mt_diagnose_clean_track()` shows you where they are),
label each fix active or stationary, and pass the resulting vector:

```{r from-speed, eval = FALSE}
library(move2)

## Per-fix step speed in m/s, via move2's helper.
v <- as.numeric(suppressWarnings(mt_speed(x, units = "m/s")))

## Smooth it so a single bad fix does not flip the label around it.
v_smooth <- stats::filter(v, rep(1/9, 9), sides = 2, circular = FALSE)

## Cut at a valley reported by mt_diagnose_clean_track() (~ 1 m/s for
## stork-class species; species- and study-specific in practice).
state <- ifelse(is.na(v_smooth) | v_smooth > 1,  "active", "stationary")

x_clean <- mt_clean_track(x, v_max = 50, state = state)
```

### From an HMM (or BCPA, or any other principled segmenter)

When you need a more rigorous split (e.g. soaring versus flapping
flight, or three states), fit a Hidden Markov Model with `momentuHMM`
or `moveHMM` and pass the decoded state sequence:

```{r from-hmm, eval = FALSE}
library(momentuHMM)

prep   <- prepData(x, type = "UTM", coordNames = c("x", "y"))
m_hmm  <- fitHMM(prep, nbStates = 3,
                 stepPar0 = c(...), anglePar0 = c(...))
state  <- viterbi(m_hmm)

x_clean <- mt_clean_track(x, v_max = 50, state = state)
```

Whatever you pass is taken as final — the package's only job is to
respect your split and clean each run.

## A worked example with synthetic data

We make up a two-state label on the package's `synthetic_tracks`
fixture (CPF_A, the densely sampled track with 23 graded outliers)
and check that `state =` returns the same number of rows and a
sensible flag pattern.

```{r worked, message = FALSE}
library(move2)
library(move2utils)

m <- read.csv(gzfile(system.file("extdata",
                                  "synthetic_tracks.csv.gz",
                                  package = "move2utils")),
              stringsAsFactors = FALSE)
m$timestamp <- as.POSIXct(m$timestamp, tz = "UTC")
mA <- mt_as_move2(m[m$individual.local.identifier == "CPF_A", ],
                  coords = c("location.long", "location.lat"),
                  time_column = "timestamp",
                  track_id_column = "individual.local.identifier",
                  crs = 4326)

## Made-up two-state label: first half "rest", second half "flight".
state <- c(rep("rest",   floor(nrow(mA) / 2)),
           rep("flight", nrow(mA) - floor(nrow(mA) / 2)))

out <- mt_clean_track(mA, v_max = 30, state = state,
                       plot = FALSE, remove = FALSE, silent = TRUE)

table(state, flagged = out$is_outlier)
```

Compare against the no-state baseline:

```{r baseline, message = FALSE}
out_pooled <- mt_clean_track(mA, v_max = 30,
                              plot = FALSE, remove = FALSE, silent = TRUE)
sum(out_pooled$is_outlier)
sum(out$is_outlier)
```

On clean ground-truth data the two should be similar. On a real
multi-behaviour track the split call usually catches outliers the
pooled call misses (in the smaller behaviour's tail) and avoids
over-flagging the body of the smaller behaviour.

Both calls use the default flag rule (`consensus =
"evidence_corroborated"`): each check's evidence is weighed on a
common scale, and a location is flagged when the combined evidence is
positive **and** either at least two checks agree or the there-and-back
check alone is overwhelmingly sure. The `state =` split changes only
*what each check sees*, not how the flags are combined. The
`error_class` column, which names the rule behind each flag, is
filled in regardless of the consensus rule. Other rules are available
via `consensus =` — including the earlier `"class_aware"`, plus
`"strict"`, `"majority"`, `"speed_trusted"`, `"weighted_evidence"`,
`"any"` and `"custom"` (see `?mt_flag_consensus`).

## Acknowledged limitations

- **Segmentation is a research problem in itself.** A naive cut at a
  density valley on log-speed can mislabel in-between states
  ("soaring" versus "active flight" on a stork). HMM-based
  segmentation with `momentuHMM` is more rigorous; manual annotation
  is the gold standard.
- **Endpoint effects.** Fixes near a state boundary are scored
  against neighbours that may belong to the other state — the path
  position check is especially sensitive to this. Smoothing the state
  label (as in the speed-threshold recipe above) reduces it.
- **Boundary outliers can slip through.** Block-shaped errors that
  straddle a state boundary may be caught by neither segment. A final
  whole-track pass with a hard `v_max` and `expand_blocks = TRUE`
  (the default) is a reasonable belt-and-braces step:

  ```{r belt, eval = FALSE}
  out <- mt_clean_track(x, v_max = 50, state = state,
                         expand_blocks = TRUE)
  ## then a final whole-track sweep with state = NULL to catch
  ## anything that straddles a boundary
  out <- mt_clean_track(out, v_max = 50, expand_blocks = TRUE)
  ```
- **Multi-individual cohorts** work transparently — if the state
  vector spans several track ids, each (track id × state run) becomes
  its own segment. State labels need not be unique across
  individuals.

## Further reading

- `vignette("OUTLIER_1_getting_started", package = "move2utils")` — the
  unified `mt_clean_track()` workflow and a brief tour of all four
  checks.
- `vignette("OUTLIER_2_diagnose_clean_track", package = "move2utils")` — the
  post-run health check.
- `vignette("OUTLIER_4_outlier_bridge", package = "move2utils")` — the path
  position check and the directional error-type classifier, for users
  who want fine control over just one check.
- `vignette("OUTLIER_5_persistence_score", package = "move2utils")` —
  multi-scale scoring of how confident each flag is; useful as a
  post-cleaning confidence filter.
- `vignette("OUTLIER_heterogeneous_error_regimes", package = "move2utils")` —
  tracks mixing sensors with different error, cleaned one sensor at a
  time.
- `vignette("OUTLIER_example_outlier_whitestork", package = "move2utils")` — a
  full narrated cleaning pipeline on a real high-frequency stork track.
- `vignette("OUTLIER_example_leo_migration", package = "move2utils")` —
  outlier detection on irregular, large-scale satellite data.
