---
title: "Migratory satellite tracking: Leo the Turkey Vulture"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Migratory satellite tracking: Leo the Turkey Vulture}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 4,
  message = TRUE
)
```

This vignette cleans a real satellite track end to end: Leo, a migratory
Turkey Vulture (*Cathartes aura*) followed from Canada to Venezuela. The
data look nothing like high-frequency GPS. Fixes arrive about three times
a day but very unevenly, the track spans a continent, and the step
lengths swing enormously between migration and the long stationary spells
on the wintering and breeding grounds. Those features are exactly what
trips up a naive outlier check, so Leo is a good test of cleaning that
respects how much time passed between locations.

## Load data

Leo's track (2007--2013) is bundled with the package. We drop empty
geometries first, as always.

```{r load, message = FALSE}
library(move2)
library(sf)
library(move2utils)

Leo <- mt_read(system.file("extdata", "Leo-65545.csv.gz",
                           package = "move2utils"))
Leo <- Leo[!st_is_empty(Leo), ]

cat(nrow(Leo), "locations over",
    round(as.numeric(diff(range(mt_time(Leo))), units = "days")),
    "days\n")

## time lag distribution
tl <- as.numeric(diff(mt_time(Leo)), units = "hours")
cat("Time lags: median", round(median(tl), 1), "h,",
    "range", round(min(tl), 1), "--", round(max(tl), 1), "h\n")
```

The gaps between fixes run from about an hour to months (the tag off, or
the bird sitting out the winter). With timing this irregular, the cleaner
has to account for elapsed time, or it will mistake a long migratory leg
for an error.

## Why elapsed time matters

If you ignore how much time passed between two locations, a 500 km
migratory step looks impossibly far compared with a day of local
foraging, and the unusual-movement check (`mt_flag_outliers()`) flags it.
Turn on `time_normalize` and the same step becomes a moderate speed --- a
vulture covering 500 km in three days is entirely normal. The comparison
makes the point:

```{r time-norm-comparison}
r_tn  <- mt_flag_outliers(Leo, time_normalize = TRUE,  plot = FALSE)
r_raw <- mt_flag_outliers(Leo, time_normalize = FALSE, plot = FALSE)

cat("With elapsed-time normalisation:   ", sum(r_tn$is_outlier), "outliers\n")
cat("Without elapsed-time normalisation:", sum(r_raw$is_outlier), "outliers\n")
```

Almost all of the extra "outliers" in the second run are false alarms:
ordinary migratory movements that only look extreme because the raw step
length ignores the days that elapsed. This is why `time_normalize = TRUE`
is the sensible choice for irregular data.

## Default detection

```{r default, fig.width = 10, fig.height = 5}
result <- mt_flag_outliers(Leo)
```

With elapsed-time normalisation and the default gap-based threshold,
Leo's track shows very few or no outliers. A soaring bird tracked by
satellite is naturally more variable than a mammal tracked by
high-frequency GPS, and the gap threshold reads that wide spread as the
normal distribution rather than chopping off its tails.

## The one-call workflow --- `mt_clean_track()`

For routine cleaning, `mt_clean_track()` runs all four checks --- path
position (`mt_flag_outliers_bridge`), there-and-back spikes
(`mt_flag_outliers_detour`), unusual movement (`mt_flag_outliers`), and
speed (`mt_flag_speed_cap`) --- under one iterative call and combines them
into a single decision. By default it weighs how strongly each check
distrusts a location on a common scale and flags the location when the
combined evidence is positive **and** either two checks agree or the
there-and-back check alone is overwhelmingly sure (`consensus =
"evidence_corroborated"`). You can switch rules via `consensus =`: the
earlier behaviour is `"class_aware"`, and `"strict"`, `"majority"`,
`"speed_trusted"`, `"weighted_evidence"`, `"any"` and `"custom"` are also
available (see `?mt_flag_consensus`).

Turkey vultures sustain flight speeds around 25 m/s. Supplying that as a
physiological cap (`v_max = 25`) lets the speed check peel any spoof- or
teleport-type clusters at their edges before the other checks run on what
survives.

```{r clean-track, fig.width = 10, fig.height = 5}
clean <- mt_clean_track(Leo, v_max = 25, plot = FALSE)
cat("kept", nrow(clean), "of", nrow(Leo), "fixes\n")
```

For long, irregular satellite tracks where you have no known top speed,
leave the cap at its data-driven default (`v_max = NULL`) --- that is the
safe call.

Whichever consensus rule you use, `mt_clean_track()` still records *why*
each location was flagged. The `error_class` column names the rule that
caught it (a clear there-and-back spike, an impossible speed, membership
of a removed block, and so on); it is computed regardless of the
consensus rule. To see the full evidence weighed on a common scale ---
the `combined_evidence` column --- run the standalone
`mt_flag_consensus(mode = "evidence_corroborated")` or
`mt_flag_consensus(mode = "weighted_evidence")` on a flagged track.

## Auto-optimised alpha

The unusual-movement check blends each step with the change in step from
one move to the next; `alpha` sets how much that change counts. For Leo's
irregular data, how much *should* it count? Let the optimiser decide:

```{r auto-alpha}
r_auto <- mt_flag_outliers(Leo, autodiff_alpha = "auto", plot = FALSE)
cat("Outliers with auto-alpha:", sum(r_auto$is_outlier), "\n")
```

## Comparison with a regular-sampling track

The contrast with regular, high-frequency sampling is instructive. The
bundled `CPF_A` synthetic track (also used in vignettes 1, 3, 4 and 5)
has regular sampling and 23 errors planted at known positions. There the
check finds genuine errors that stand out clearly against an otherwise
tight distribution:

```{r regular-comparison}
path <- system.file("extdata/synthetic_tracks.csv.gz",
                    package = "move2utils")
tracks <- mt_read(path)
cpf_a  <- filter_track_data(tracks, .track_id = "CPF_A")

r_cpf <- mt_flag_outliers(cpf_a, plot = FALSE)
cat("CPF_A (regular sampling, 23 planted outliers):",
    sum(r_cpf$is_outlier), "flagged of", nrow(cpf_a), "fixes\n")
cat("Leo (vulture, satellite):",
    sum(result$is_outlier), "flagged of", nrow(Leo), "fixes\n")
```

The check adapts to the data. Regular high-frequency sampling gives a
tight distribution where planted errors jump out; irregular satellite
data has wide natural variability that the gap threshold respects rather
than punishes.

## Multi-scale persistence annotation

`mt_persistence_score()` takes any flagged output and adds a confidence
score: at each of `scales = c(2, 4, 8)` (the validation scales) it asks
whether a flagged location still looks anomalous when viewed over a wider
time window. Scores run from 1 (flagged only at the native resolution) to
4 (flagged at every scale). It does not touch `is_outlier` --- it only
annotates.

```{r persistence}
annotated <- mt_persistence_score(result, silent = TRUE)
flagged <- which(annotated$is_outlier)
cat("Persistence score distribution among flagged fixes:\n")
print(table(annotated$persistence_count[flagged]))
```

See `vignette("OUTLIER_5_persistence_score", package = "move2utils")` for
the empirical class-conditional analysis on synthetic CPF data.

## Further reading

- `vignette("OUTLIER_1_getting_started", package = "move2utils")` — the
  unified `mt_clean_track()` workflow and a brief tour of all four
  checks.
- `vignette("OUTLIER_2_diagnose_clean_track", package = "move2utils")` —
  the after-cleaning health check.
- `vignette("OUTLIER_3_state_conditional", package = "move2utils")` —
  cleaning different behaviours separately, the recipe to reach for when
  the diagnostic flags two behavioural states.
- `vignette("OUTLIER_4_outlier_bridge", package = "move2utils")` — the
  path-position check and the kinds of error its directional variant
  distinguishes; for fine control over a single check.
- `vignette("OUTLIER_5_persistence_score", package = "move2utils")` —
  scoring how confident each flag is; useful as a post-cleaning filter.
- `vignette("OUTLIER_heterogeneous_error_regimes", package = "move2utils")` —
  tracks mixing sensors with different error, cleaned one sensor at a
  time.
- `vignette("OUTLIER_example_outlier_whitestork", package = "move2utils")` —
  a full narrated cleaning on a real high-frequency stork track.
