---
title: "Multi-scale persistence score for outlier flags"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-scale persistence score for outlier flags}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 9,
  fig.height = 5,
  message  = TRUE
)
library(move2utils)
library(move2)
library(sf)
```

## What this vignette covers

Once `mt_clean_track()` has flagged a fix, the natural next question
is how confident you should be that the fix is really an error.
`mt_persistence_score()` gives you a number for that. It re-checks
each flagged fix at wider time gaps: if a fix looks wrong compared to
its immediate neighbours, *and* still looks wrong compared to the
fixes 2, 4, and 8 steps away, that is stronger evidence than a fix
that only looks odd at the finest scale.

The result is a per-fix confidence score from 1 to 4 (with the default
settings). You decide whether to drop the low-confidence flags or keep
them all. The function never changes `is_outlier` itself.

It works on the output of any flagger in the package —
`mt_clean_track()` (the full cleaner), `mt_flag_outliers_bridge()`
(path position), `mt_flag_outliers_detour()` (there-and-back),
`mt_flag_outliers()` (unusual movement), `mt_flag_speed_cap()`
(speed), `mt_sequential_outliers()`, and `mt_combined_outliers()`. It
does not care *how* a fix was flagged, only whether it still looks out
of place when you compare across wider gaps.

The worked examples below show that the score is most useful when you
filter it according to the cleaner's `error_class` column, rather than
applying one rule to every flag. We show what that looks like in
practice.

## How it works — the idea first

For each flagged fix the function asks: if I compare this fix to the
one 2 steps away instead of 1 step away — and again at 4 steps and 8
steps — does it still look like an outlier?

The "looks like an outlier" test at each scale uses the same kind of
joint distribution that the per-fix checks use, just on windowed step
lengths and turn angles instead of single-step ones. The reference at
each scale is built from *every* interior fix of the track, not only
the flagged ones, so each candidate is scored against the track's own
geometry at that scale rather than against an outside assumption.

A fix that trips only at the native single-step resolution gets score
1. A fix that also trips at the 2-step, 4-step, and 8-step windows
gets score 4. The higher the score, the more robust the anomaly: it
survives looking at the track more coarsely.

### The formal picture (for the curious)

For each flagged fix $i$ and each validation scale $k \in
\{2, 4, 8\}$ (the default), the function computes:

- $\text{step}_k^\text{in}(i)$: distance from $x_{i-k}$ to $x_i$
- $\text{step}_k^\text{out}(i)$: distance from $x_i$ to $x_{i+k}$
- $\text{turn}_k(i)$: turn angle at $x_i$ between the two long arms

The reference distribution at scale $k$ is built from the same
quantities computed at *every* interior fix of the track (the
candidate is not removed). A 2-D histogram of
$(\log\text{step}_k, \text{turn}_k)$ gives the joint density; a
gap-on-($-\log$ probability) threshold decides whether each candidate
is flagged at scale $k$.

The persistence score is

$$
p(i) = 1 + \sum_{k \in \text{scales}} \mathbb{1}\bigl[i \text{ flagged at scale } k\bigr],
$$

so $p(i) \in \{1, 2, 3, 4\}$ for the default `scales = c(2, 4, 8)`.

#### Why no thinning?

A natural-sounding alternative is multi-scale voting: thin the track
at each scale, run a detector on each thinned subset, then count
votes. That has a structural index-parity problem. A fix at original
index 7, with `scales = c(1, 2, 4, 8)`, is in the thinned subset only
at scale 1; it can vote at most once, no matter how anomalous it looks
at the wider scales. The persistence score sidesteps this by
evaluating *every* flagged fix at *every* validation scale.

## Worked example

Load the bundled synthetic data and pull out the spike-contaminated
`CPF_A` track. We also read the ground truth, which records the index
of each injected error.

```{r cpf-data, message = FALSE}
path   <- system.file("extdata/synthetic_tracks.csv.gz", package = "move2utils")
tracks <- mt_read(path)
cpf_a  <- filter_track_data(tracks, .track_id = "CPF_A")

gt      <- readRDS(system.file("extdata", "synthetic_ground_truth.rds",
                               package = "move2utils"))
truth_a <- gt[["CPF_A"]]$index
```

Run the cleaner on `CPF_A` and annotate the result with the
persistence score:

```{r cascade-and-score}
clean <- mt_clean_track(cpf_a, plot = FALSE, remove = FALSE)
ann <- mt_persistence_score(clean, silent = TRUE)

flagged <- which(ann$is_outlier)
cat("Cleaner flagged:", length(flagged), "fixes\n")
cat("Persistence score distribution:\n")
print(table(ann$persistence_count[flagged]))
```

The printed distribution shows that most flags persist at the maximum
score (`p = 4`), as expected for spike-class outliers whose anomaly
survives every coarsening.

## The class-conditional finding

The empirical question is whether persistence usefully separates true
positives from false positives within the cleaner's flag set. The
answer depends on which `error_class` the cleaner assigned.

`mt_clean_track()` attaches one of six error classes to each flag:
`geometric_spike`, `consensus`, `state_anomaly`,
`kinematic_confluence`, `block`, `physiological`. Each class records a
different combination of the four checks that fired, and each has a
different empirical relationship with persistence. The `error_class`
column is computed the same way whatever consensus rule decided the
flags, so the pattern below holds under the current default
(`evidence_corroborated`) just as it did under the earlier
`class_aware` rule.

Pooling across all five CPF synthetic tracks (`CPF_A` through
`CPF_F`), and asking, for each error class, what fraction of true
positives versus false positives persist at $\geq 3$ scales:

| `error_class` | $n$ TPs | $n$ FPs | % TPs at $p \geq 3$ | % FPs at $p \geq 3$ | TP$-$FP gap |
|---|---|---|---|---|---|
| `geometric_spike` | 76 | 0 | 51% | (no FPs) | (class-pure on synthetic) |
| `state_anomaly` | 24 | 23 | 96% | 57% | **+39.3 pp** |
| `consensus` | 34 | 9 | 82% | 44% | **+37.9 pp** |
| `kinematic_confluence` | 7 | 7 | 71% | 86% | $-$14.3 pp (small $n$) |

Three points:

1. The `geometric_spike` class is **empirically pure on the
   synthetic**: the cleaner produces no false positives here, so
   persistence has nothing to filter.
2. The `state_anomaly` and `consensus` classes show a clean
   **+37–39 pp** gap: true positives persist far more often than false
   positives. A persistence-based filter on these classes
   substantively improves precision.
3. The `kinematic_confluence` class rests on a small sample (7 TPs / 7
   FPs, concentrated on the multi-state `CPF_F` track), and the signal
   direction is uncertain.

## Recommended usage pattern

Use the helper as **annotation only** by default; never let it change
`is_outlier` automatically. Where you do want to filter, gate the
filter on `error_class` so it acts only where persistence is
discriminative:

```{r class-aware-filter}
clean <- mt_clean_track(cpf_a, plot = FALSE, remove = FALSE)
ann <- mt_persistence_score(clean, silent = TRUE)

## Drop low-persistence flags only in the classes where persistence
## is empirically discriminative.
demote <- ann$is_outlier &
            ann$error_class %in% c("state_anomaly", "consensus") &
            ann$persistence_count < 3

cat("Cleaner flags:           ", sum(ann$is_outlier), "\n")
cat("Demoted by class filter: ", sum(demote), "\n")
cat("Final flagged after filter:",
    sum(ann$is_outlier & !demote), "\n")
```

## When not to apply the persistence filter

The score loses its discriminative power in two cases:

- **Halo-style outliers.** Repeated wandering returns to a fixed
  location have anomalies that *average out* over wider windows, so
  their persistence score is structurally low even when the fix is
  genuinely an error. The cleaner usually files these as
  `geometric_spike` (where the class is already pure and no filtering
  is needed). But on a detector's raw output, which carries no
  `error_class` column, applying a persistence filter to halo data
  will preferentially remove true positives.

- **Bridge or detour standalone output.** On raw path-position
  (`mt_flag_outliers_bridge()`) or there-and-back
  (`mt_flag_outliers_detour()`) flags, false positives *persist more*
  than true positives, so a persistence filter would hurt precision
  rather than help. Both checks flag on local-window geometry, so
  their true positives include scale-dependent cases (halo), while
  their false positives sit at structurally complex track regions
  where any window-based score fires consistently.

The safe pattern is therefore: **annotate, then filter by
`error_class`**. Without an error-class column, treat the persistence
score as a confidence column you inspect, not an automatic filter.

## Custom scales

The default `scales = c(2, 4, 8)` suits typical animal-tracking data.
Adjust it if your sampling rate or contamination geometry warrants:

```{r custom-scales, eval = FALSE}
## Denser scale ladder for high-frequency GPS:
ann <- mt_persistence_score(clean, scales = c(2L, 3L, 5L, 8L, 13L))

## Coarser-only validation for sparse Argos data:
ann <- mt_persistence_score(clean, scales = c(4L, 8L, 16L))
```

Each scale must be $\geq 2$; scale 1 is the original flagger's
resolution and is already counted in the score's $+1$ constant.

## Relationship to the retired `mt_flag_outliers_multiscale`

`mt_persistence_score()` replaces the previous
`mt_flag_outliers_multiscale()` (retired in v0.3). The predecessor
thinned the track at multiple scales and voted; the new helper does
not thin and instead computes a windowed view at every fix. This
removes the index-parity bias of the old design and makes the score
detector-agnostic. See the v0.3 NEWS entry for the empirical
motivation behind the change.

## Further reading

- `vignette("OUTLIER_1_getting_started", package = "move2utils")` — the
  unified `mt_clean_track()` workflow and a brief tour of all four
  checks.
- `vignette("OUTLIER_2_diagnose_clean_track", package = "move2utils")` — the
  post-run health check.
- `vignette("OUTLIER_3_state_conditional", package = "move2utils")` — when the
  diagnostic flags two behaviours, the recipe for cleaning each
  behavioural state separately.
- `vignette("OUTLIER_4_outlier_bridge", package = "move2utils")` — the
  path-position check and the directional error-morphology classifier,
  for users who want fine-grained control over just one detector.
- `vignette("OUTLIER_heterogeneous_error_regimes", package = "move2utils")` —
  cleaning tracks that mix sensors with different error, one sensor at
  a time.
- `vignette("OUTLIER_example_outlier_whitestork", package = "move2utils")` — a
  full narrated cleaning on a real high-frequency stork track.
- `vignette("OUTLIER_example_leo_migration", package = "move2utils")` —
  cleaning irregular, large-scale satellite data.
</content>
</invoke>
