---
title: "Troubleshooting: when mt_clean_track() doesn't do what you want"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Troubleshooting: when mt_clean_track() doesn't do what you want}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 4,
  message = TRUE
)
library(move2utils)
```

`mt_clean_track()` works well on most tracks, but no automatic cleaner is
right every time — and because there is no ground truth for outliers,
"right" is partly your judgement. This guide maps the common symptoms to
their usual causes and remedies.

Whatever the symptom, start the same way: **plot the result, and run the
health check** `mt_diagnose_clean_track()` (see
`vignette("OUTLIER_2_diagnose_clean_track", package = "move2utils")`). It
tells apart "the track was dirty" from "the cleaner was misled", and most
of the remedies below follow directly from what it shows.

| Symptom | Usual cause | Remedy |
|---|---|---|
| Removed too much | animal has two behaviours | clean each state separately |
| Removed too much | coarse / irregular sampling | clean at native resolution; mind the scale |
| Removed too much | inferred speed cap too low | supply `v_max` or `mass` + `mode` |
| Removed too much | mixed sensors / error regimes | clean one regime per call |
| Missed obvious errors | coherent spoof / jam block | supply `v_max` to engage block removal |
| Missed obvious errors | rare modest spike on a slow species | inspect by hand; try a more sensitive rule |
| Missed obvious errors | whole-track shift / off-site cluster | check distances by hand (step-level checks can't see it) |
| Error / crash | empty geometries, duplicate or unsorted times, wrong CRS | pre-process (below) |

## It removed too much

### The animal has two behaviours

The most common cause. An animal that both rests and commutes (or
migrates) has two very different kinds of step. Lumped together, the
slower behaviour's normal steps can look anomalous against the faster
one, and get flagged. `mt_diagnose_clean_track()` flags this as bimodal
behaviour. The remedy is to clean each behaviour separately, by passing a
behaviour column to `state =`:

```{r state, eval = FALSE}
clean <- mt_clean_track(track, state = "behaviour")
```

See `vignette("OUTLIER_3_state_conditional", package = "move2utils")`.

### The sampling is coarse or irregular

Cleaning is not scale-invariant: the same track sampled every few seconds
and every few hours can clean differently, because coarse sampling can
make ordinary rest-and-travel structure look bimodal. If you downsampled
the track, try cleaning at the original resolution; if the tag itself
samples coarsely, treat a raised flag rate with the state-conditional
remedy above rather than accepting it.

### The inferred speed cap is too low

With no `v_max`, the cleaner infers a speed cap from the track. On a
fast-moving species this can land inside the animal's real speed range,
so the block-removal step over-fires. Give it the animal's real top speed
(or a body-mass estimate) and check the candidates first:

```{r cap, eval = FALSE}
mt_suggest_speed_cap(track, mass = 5, mode = "flying")  # inspect, then:
clean <- mt_clean_track(track, v_max = 30)
```

### The track mixes sensors with different error

A track that combines, say, GPS and Argos Doppler fixes mixes two very
different error regimes, and a single cleaning pass cannot serve both.
Clean one regime at a time. Network-side geolocation streams (Sigfox-geo,
some LoRa) are not position estimates at all and should be excluded
entirely, not cleaned. See
`vignette("OUTLIER_heterogeneous_error_regimes", package = "move2utils")`.

## It missed obvious errors

### A coherent block of fake locations (spoof or jam)

A run of fabricated locations (a GPS spoof, or a jamming burst) can be
internally consistent — its individual steps are small — so the per-fix
checks only see its edges. Giving the cleaner the animal's real top
speed, via `v_max` or `mass` + `mode`, hands it a firm connectivity
ceiling so it can recover the whole block, not just the seam.

The package ships a fully synthetic demonstration track that carries one
of every gross-error class — including a coherent spoof block and a
jamming burst — with a ground-truth `error_type` column you can score
against:

```{r spoof-setup}
source(system.file("extdata", "make_demo_track.R", package = "move2utils"))
track <- make_demo_track()
## the same pre-processing as "It errored or crashed", below
track <- track[!sf::st_is_empty(track), ]
track <- move2::mt_filter_unique(track, criterion = "first")
```

```{r spoof-clean}
clean <- mt_clean_track(track, mass = 0.5, mode = "flying",
                        remove = FALSE, silent = TRUE)

## how much of each fabricated block was recovered?
table(error_type = clean$error_type, flagged = clean$is_outlier)[
  c("spoof", "jam"), ]
```

Both blocks are recovered almost in full. Supplying the cap is the
robust choice for coherent blocks: when a block sits far enough off-route
that only its boundary steps look fast, the cap is exactly the ceiling
the block-removal step needs to isolate it — and supplying one can only
help, never hurt, block recovery.

### A rare, modest spike on a slow-moving animal

A single out-and-back excursion on a slow species can be small enough to
sit within ordinary GPS error — there is simply no reliable signal to
separate it from good data, and no automatic method can recover what
isn't there. Do not chase these blindly. If you suspect them, inspect the
track, try the more sensitive `consensus = "weighted_evidence"` rule (see
the next section), and remove the rest by hand.

### A whole-track shift or an off-site cluster

The four checks are step-level: they compare each location to its
neighbours. A whole-track offset, or a tight cluster of fixes parked at
one wrong place, has no unusual *step* and so slips through. Check it
directly — for example, the distance of each location from the track's
centre — and remove by hand.

## It errored or crashed

Most errors come from the track not being in the expected shape. Apply
the standard pre-processing first:

```{r preprocess, eval = FALSE}
library(move2); library(sf); library(dplyr)
track <- track[!st_is_empty(track), ]          # drop empty geometries
track <- mt_filter_unique(track, "first")      # drop duplicate timestamps
track <- arrange(track, mt_track_id(track), mt_time(track))  # sort in time
```

`mt_clean_track()` auto-projects longitude/latitude internally, so you do
not need to project first. If an error persists after pre-processing,
please report it.

## Choosing a different decision rule

The default rule, `evidence_corroborated`, flags a location when the
checks' combined evidence is positive and either two checks agree or the
there-and-back check alone is overwhelming. Two alternatives are worth
knowing, set via `consensus =`:

- `"class_aware"` — the previous default. Stricter about requiring
  corroboration, so slightly better on coherent clusters; use it if the
  default removes a near-bulk point you wanted kept.
- `"weighted_evidence"` — more sensitive: it flags on positive combined
  evidence alone, without the corroboration safeguard. Use it when the
  default misses a conspicuous point, accepting a few more false alarms.

```{r consensus, eval = FALSE}
clean <- mt_clean_track(track, consensus = "weighted_evidence")
```

For fine control, the standalone `mt_flag_consensus()` returns a
`combined_evidence` column — a single number per location summarising how
strongly the checks distrust it. Sorting by it, or thresholding it
yourself, gives you a one-dial sensitivity control without changing any
detector.

## Still stuck?

Each check is available on its own — `mt_flag_outliers_bridge()` (path
position), `mt_flag_outliers_detour()` (there-and-back),
`mt_flag_outliers()` (unusual movement), `mt_flag_speed_cap()` (speed) —
so you can inspect one signal at a time, or build a pipeline of your own.
The `OUTLIER_example_*` vignettes show full cleanings narrated on real
tracks, which is often the quickest way to see how the pieces fit your
own data.
