---
title: "Bridge-residual outlier detection and error-morphology classification"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bridge-residual outlier detection and error-morphology classification}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 5,
  message = TRUE
)
```

This vignette is about the **path-position check**,
`mt_flag_outliers_bridge()`. Use it on its own when you want to find
GPS fixes that sit in the wrong place — a single location that jumps
far from where its neighbours are, then comes straight back. It is the
geometric one of the four checks that `mt_clean_track()` combines;
running it alone lets you look at just that one signal, or build your
own pipeline on top of it.

We cover what the check does, the directional variant that also tells
you *what kind* of error you found, and how to read the columns it adds.

If you are new to the package, start with `mt_clean_track()` instead —
it runs this check for you alongside three others and decides what to
flag. Come back here when you want fine control over this one signal.
The unusual-movement check `mt_flag_outliers()` is covered in
`vignette("OUTLIER_1_getting_started", package = "move2utils")`. Both
checks share the same threshold machinery but target different kinds of
error, and can be combined.

## The idea

If a GPS fix does not belong, the simplest test is: given where the
animal was just before and just after, where would you *expect* the
fix in between to be? If the actual fix is far from that expectation —
further than the time gap on either side could account for — it is
probably wrong.

That is the bridge. For each fix we draw a line (the "bridge") between
its two neighbours in time, weighted by how much time sits on either
side, and measure how far the actual fix lies from that line. A small
distance means the fix is about where it should be; a large one means
something is off.

The key point is that the *width* of the bridge — how much wandering is
plausible over a given time gap — depends only on the timestamps, never
on the fix's own position. A bad fix therefore cannot inflate the
tolerance that would have flagged it. Methods that instead scale by a
locally-estimated variance suffer from this "leverage problem", because
a bad fix pollutes its own scale.

### The formula (for the curious)

For each fix *i* with projected coordinates **p**_i and time *t*_i, its
neighbours *i-1* and *i+1* define a Brownian-bridge expectation

> m_i = α **p**_{i-1} + (1 − α) **p**_{i+1},   α = Δt₂ / (Δt₁ + Δt₂)

where Δt₁ = t_i − t_{i-1} and Δt₂ = t_{i+1} − t_i.

The residual is **r**_i = **p**_i − m_i. It is scaled by the bridge's
own standard-deviation width,

> w_i = √(Δt₁ · Δt₂ / (Δt₁ + Δt₂)).

The **bridge score** is η_i = ||**r**_i|| / w_i. As noted above, w_i
depends only on timestamps, never on positions, so an outlier cannot
inflate its own denominator.

### Three methods

- **`method = "isotropic"`** uses the plain residual magnitude.
  General-purpose; answers "is this fix far from where it should be?".
- **`method = "directional"`** splits the residual along the local
  travel direction into a parallel part (η_para, along-track) and a
  perpendicular part (η_perp, across-track). It flags on η_perp alone,
  but the value is in the (η_para, η_perp) pair — it tells you *what
  kind* of error each flag is.
- **`method = "combined"`** (default) applies the threshold to both
  scores and flags a fix if either one trips.

The check borrows its bridge-mean construction from the dynamic
Brownian bridge (dBBMM) and its perpendicular split from the dynamic
bivariate Gaussian bridge (dBGB), but deliberately does *not* use their
variance-estimation step — that is what avoids the leverage problem.
See `mt_dbbmm_variance()` / `mt_dbgb_variance()` for the full dynamic
models.

## Requirements

- The function accepts longitude/latitude or projected input; it
  projects internally to a local AEQD for the distance math and returns
  the result in your original CRS.
- **Enough context.** With only a handful of locations the bridge
  neighbours are undefined; in practice you need at least a few dozen
  locations for the threshold estimator to be stable.

## A worked example

`inst/extdata/synthetic_tracks.csv.gz` contains central-place-forager
tracks with ground-truth outliers for validation. CPF_A has 23 injected
outliers, CPF_B is clean, and CPF_C has 4.

```{r load, message = FALSE}
library(move2)
library(sf)
library(move2utils)

path <- system.file("extdata/synthetic_tracks.csv.gz",
                     package = "move2utils")
tracks <- mt_read(path)
tracks <- tracks[!st_is_empty(tracks), ]

cpfA <- tracks[mt_track_id(tracks) == "CPF_A", ]
cpfA_p <- st_transform(cpfA, mt_aeqd_crs(cpfA))
nrow(cpfA_p)
```

## Default run (combined method, entropy threshold)

```{r run-default}
res <- mt_flag_outliers_bridge(cpfA_p, plot = FALSE)
table(res$is_outlier)
```

```{r flag-map, fig.width = 7, fig.height = 6, fig.cap = "CPF_A synthetic track in projected coordinates. Grey line traces the full track in order; firebrick circles mark the fixes mt_flag_outliers_bridge() flagged under the default combined-entropy configuration. Flagged points sit visibly off the otherwise smooth central-place-forager geometry."}
coords <- sf::st_coordinates(res)
par(mar = c(4, 4, 3, 1))
plot(coords, type = "l", col = "grey80", asp = 1,
     xlab = "x (m)", ylab = "y (m)",
     main = sprintf("CPF_A — %d flags (combined, entropy)", sum(res$is_outlier)))
points(coords[!res$is_outlier, ], pch = 16, cex = 0.25, col = "grey50")
points(coords[res$is_outlier, ], pch = 1, cex = 1.6,
       col = "firebrick", lwd = 1.4)
legend("topright", pch = c(16, 1), col = c("grey50", "firebrick"),
       legend = c("kept", "flagged"), bty = "n")
```

The columns added to the input are

| column | meaning |
|---|---|
| `bridge_residual` | residual magnitude ||**r**_i|| |
| `bridge_width`    | bridge-width normalisation w_i |
| `bridge_eta`      | score η_i = residual / width (scalar / isotropic) |
| `bridge_percentile` | empirical quantile of η within the track |
| `bridge_iteration`  | iteration on which the fix was flagged (0 = not flagged) |
| `is_outlier`        | the flag |

The `"entropy"` threshold (default) is the deepest valley in the
kernel-density estimate of log(η). On clean data the density has a
single hump, there is no valley, and the function returns zero flags —
a **no-op guarantee** that makes it safe to apply to any track. The
`"gap"` threshold is more sensitive: it uses a broken-stick null model
plus tail-decay inflection, and will often pick up borderline points
that entropy leaves alone.

```{r run-gap}
res_gap <- mt_flag_outliers_bridge(
  cpfA_p,
  threshold_type = "gap",
  plot = FALSE
)
table(res_gap$is_outlier)
```

Use `"gap"` when you want sensitivity and can tolerate a few false
positives; use `"entropy"` (the default) when false positives cost more
than missed outliers, or when the track might already be clean.

## Directional variant — what kind of error?

```{r run-dbgb}
res_d <- mt_flag_outliers_bridge(
  cpfA_p,
  method = "directional",
  plot   = FALSE
)
grep("bridge", names(res_d), value = TRUE)
```

Two extra columns appear:

- `bridge_eta_para` — residual along the local travel direction
- `bridge_eta_perp` — residual across it

Flags go on `bridge_eta_perp` (the across-track signal, usually the
clearer mark of genuine measurement error). But the useful thing is the
*pair* of values for each flag.

### Reading the (η_para, η_perp) plane

```{r dbgb-scatter}
flags <- res_d[res_d$is_outlier, ]
n_flag <- nrow(flags)

plot(log10(flags$bridge_eta_para + 1),
     log10(flags$bridge_eta_perp + 1),
     xlab = expression(log[10](eta["para"] + 1)),
     ylab = expression(log[10](eta["perp"] + 1)),
     pch = 16, col = "firebrick", cex = 1.2,
     main = sprintf("error morphology on %d flagged points", n_flag))
abline(0, 1, lty = 2, col = "grey50")
```

How to read the plane:

- **Near the diagonal** — η_para ≈ η_perp. Scatter in all directions;
  classic jitter-type error.
- **Below the diagonal** (high η_para, low η_perp) — along-track jumps.
  Ghost reports, clock glitches, GNSS fold-back.
- **Above the diagonal** (low η_para, high η_perp) — across-track drift.
  Typical of multipath, reflective environments, or constellation
  changes that bias one axis.

This is why the isotropic and directional flag sets can look quite
different on real data even though both come from the same residual: the
isotropic score answers *how far*, the directional split answers *which
way*. Together they separate error types that one check would lump.

## Combining bridge with the other checks

`mt_flag_outliers_bridge()` is one of four checks in `move2utils`; the
others are the unusual-movement check `mt_flag_outliers()`
(probability-based), the there-and-back check `mt_flag_outliers_detour()`
(geometric, time-insensitive path-vs-displacement ratio), and the speed
check `mt_flag_speed_cap()` (step-level physiological cap). They are
complementary:

- `mt_flag_outliers()` is probability-based; it does best on
  multi-state, behaviourally mixed tracks where the gap-aware
  auto-difference carries the signal.
- `mt_flag_outliers_bridge()` is geometric and leverage-immune; it does
  best on long clean tracks and on drift or spoofing.
- `mt_flag_outliers_detour()` is time-insensitive and scale-invariant;
  it catches there-and-back spikes at sparse sampling, where the
  bridge check's width-scaling loses sensitivity.
- `mt_flag_speed_cap()` is step-level; it catches the edges of coherent
  multi-fix error blocks that per-fix checks cannot reach.

The unified `mt_clean_track()` runs all four for you. A simple
standalone pipeline can instead flag the union (or the intersection,
depending on your cost function) of two of them; the example below pairs
bridge with the unusual-movement check, the most common combination for
moderately-sampled tracks:

```{r compose, eval = FALSE}
prob   <- mt_flag_outliers(cpfA_p)
bridge <- mt_flag_outliers_bridge(cpfA_p, method = "directional", plot = FALSE)

both_flags <- prob$is_outlier | bridge$is_outlier
cat("probability:", sum(prob$is_outlier), "\n")
cat("bridge     :", sum(bridge$is_outlier), "\n")
cat("union      :", sum(both_flags), "\n")
```

For routine use, `mt_clean_track()` does this composition for you and
adds the speed cap, a rule that combines the four checks into one
decision, and a block-expansion pass that catches coherent multi-fix
clusters that per-fix scoring cannot resolve. By default the combine
rule is `consensus = "evidence_corroborated"`: each check's evidence is
weighed on a common scale, and a location is flagged when the combined
evidence is positive **and** either at least two checks agree or the
there-and-back check alone is overwhelmingly sure. You can pick another
rule via `consensus =` — `"class_aware"` (the earlier default),
`"strict"`, `"majority"`, `"speed_trusted"`, `"weighted_evidence"`,
`"any"` and `"custom"` (see `?mt_flag_consensus`). Supply `v_max` when
species biology gives you a physiological cap — that lets the pipeline
peel spoof- or teleport-class clusters at their edges before the bridge
and unusual-movement checks run on what survives.

```{r clean-track-link, eval = FALSE}
clean <- mt_clean_track(cpfA_p)
clean <- mt_clean_track(eagle_track, v_max = 30)  # eagle ~30 m/s ceiling
```

## Iterative refinement

A single pass can miss outliers that were themselves used as neighbours
by other outliers. `iterations = n` re-runs the check on the
currently-clean subset up to `n` times (default 3), stopping early once
nothing new is flagged. Raise it when your tracks have long runs of
consecutive bad fixes.

## Disabling neighbour smearing

Next to a real outlier, its two immediate neighbours can look anomalous
too, because one of their bridge expectations rests on a corrupted
point. `dedup_neighbours = TRUE` (default) suppresses this
neighbour-smearing with a small peak-picking pass. Turn it off only if
you suspect it is masking real clustered errors.

## Further reading

- `vignette("OUTLIER_1_getting_started", package = "move2utils")` — the
  unified `mt_clean_track()` workflow and a brief tour of all four
  checks.
- `vignette("OUTLIER_2_diagnose_clean_track", package = "move2utils")` —
  the post-run health check.
- `vignette("OUTLIER_3_state_conditional", package = "move2utils")` —
  when the diagnostic flags two behaviours, the recipe for cleaning each
  behavioural state separately.
- `vignette("OUTLIER_5_persistence_score", package = "move2utils")` —
  multi-scale annotation that scores how confident each flag is; useful
  as a post-cleaning confidence filter.
- `vignette("OUTLIER_heterogeneous_error_regimes", package = "move2utils")` —
  cleaning tracks that mix sensors with different error, one sensor at a
  time.
- `vignette("OUTLIER_example_outlier_whitestork", package = "move2utils")` —
  a full narrated cleaning on a real high-frequency stork track.
- `vignette("OUTLIER_example_leo_migration", package = "move2utils")` —
  cleaning irregular, large-scale satellite data.
- `?mt_flag_outliers_bridge` — full argument reference and tuning notes.
