---
title: "Diagnosing an mt_clean_track run"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Diagnosing an mt_clean_track run}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 9,
  fig.height = 6,
  message  = TRUE
)
```

`mt_clean_track()` gives you a result, but did it actually work? A high
flag rate can mean the track was dirty — or it can mean the cleaner was
fooled by normal behaviour. `mt_diagnose_clean_track()` is the post-run
health check that tells the two apart. It draws six panels, much like the
diagnostic plots of a fitted model, and each panel targets one way a run
can go wrong and points to a specific remedy. It does not re-run the
checks; it reads the flag columns and the step speeds and renders one
figure you can eyeball.

```{r load, message = FALSE}
library(move2)
library(sf)
library(move2utils)
```

## A run that needs attention — Pettstadt1, a juvenile white stork

Pettstadt1 is a juvenile white stork (~3 kg) with 48k GPS fixes spanning
two summers and one autumn migration. We clean it, then run the
diagnostic on the result.

```{r pet-clean}
pet <- mt_read(system.file("extdata/Pettstadt1-14053.csv.gz",
                              package = "move2utils"))
pet <- pet[!st_is_empty(pet), ]
pet <- mt_filter_unique(pet, criterion = "first")
pet <- dplyr::arrange(pet, mt_track_id(pet), mt_time(pet))
res_pet <- mt_clean_track(pet, mass = 3, mode = "flying",
                            plot = FALSE, remove = FALSE,
                            silent = TRUE,
                            max_flag_fraction = 0.5)
cat(sprintf("Pettstadt1: %d / %d flagged (%.2f%%)\n",
            sum(res_pet$is_outlier), nrow(res_pet),
            100 * mean(res_pet$is_outlier)))
```

The flag fraction is high — more than genuine GPS errors would produce on
a stork. Time to ask the diagnostic what happened.

```{r pet-diagnose, fig.width = 11, fig.height = 7}
diag <- mt_diagnose_clean_track(res_pet)
```

The function prints the concerns it found and renders six panels.

### Reading the panels

**Panel 1 — log-speed density with detected modes.** The grey curve is
the distribution of step speeds on a log scale. The blue dashed lines
mark the substantive modes the package found (a mode counts when its
basin holds at least 2 % of fixes *and* its peak density is at least 5 %
of the tallest peak). The red line is the `v_max` that was used. A
healthy track shows a single dominant mode with a thin upper tail. On
Pettstadt1 you instead see two or three modes — sleeping or perched at
the low end, flight further right. Several substantive modes is the
*bimodal-behaviour signature*: each per-location check measures a fix
against one overall distribution, so fixes from one mode can fall in the
tail of the other and look like outliers.

**Panel 2 — flag rate vs time.** Rolling 7-day flag rate along the
timeline. Healthy tracks read flat, near 0 %, with isolated narrow spikes
only where the bird had a real GPS noise event. On Pettstadt1 you see a
sustained band of elevated flag rate through the autumn migration months.
Sustained is not what errors look like — sustained means the bird was
doing something the checks were not calibrated for.

**Panel 3 — per-detector activity.** Counts of fixes flagged by each
combination of checks, summed across all iterations. The headline
categories come from the `error_class` column, which `mt_clean_track()`
fills the same way no matter which consensus rule you run, so this panel
reads the same under the default `evidence_corroborated` rule as under
any other. The categories are `consensus` (≥3 of the 4 checks agree),
`geometric_spike` (path-position **and** there-and-back), `state_anomaly`
((path-position or there-and-back) **and** speed), and
`kinematic_confluence` ((path-position or there-and-back) **and**
unusual-movement). The single-check bars — path-position-only,
there-and-back-only, unusual-movement-only, speed-only — are fixes one
check flagged that the combined decision did not promote to an outlier.
That is useful: a tall single-check bar tells you which check is firing
noisily on this track. On Pettstadt1 expect the path-position and
unusual-movement checks to dominate (`kinematic_confluence`):
path-position says "this fix sits far from a straight line between its
neighbours" and unusual-movement says "this fix's joint speed-and-turn
signature lies in the tail of the distribution". Both are correct
*relative to the resting baseline*, and both are wrong as errors during
migration.

**Panel 4 — cumulative flagging by iteration.** A rapid plateau in 2–4
iterations is healthy — the iteration loop has settled on a stable set of
flags. Linear growth that never plateaus, especially if the run hits the
`flag_fraction_exceeded` abort, is the *non-converging signature*: each
pass peels off another batch as the working set keeps shifting. There is
no fixed point because the root mismatch — a bimodal distribution against
single-distribution thresholds — cannot be resolved by iterating.

**Panel 5 — consecutive-flag run lengths.** Healthy: a tall length-1 bar
and a short tail. The orange and red bins ("5–9", "10+") count *runs* of
consecutively flagged fixes. Long runs are not what discrete GPS errors
look like; they are what happens when the bird flies through a region the
checks treat as anomalous for hours at a time.

The notes that `mt_diagnose_clean_track()` prints summarise these patterns
in plain language and point to the remedy.

## A clean run for contrast — CPF_B (synthetic, no truth outliers)

The bundled CPF_B track is documented as clean (no injected outliers), so
a healthy run and its diagnostic should look very different from
Pettstadt1.

```{r cpf-clean, fig.width = 11, fig.height = 7}
syn <- mt_read(system.file("extdata/synthetic_tracks.csv.gz",
                             package = "move2utils"))
syn <- syn[!st_is_empty(syn), ]
m_B <- syn[mt_track_id(syn) == "CPF_B", ]
res_B <- mt_clean_track(m_B, plot = FALSE, remove = FALSE, silent = TRUE)
cat(sprintf("CPF_B: %d / %d flagged (%.4f%%)\n",
            sum(res_B$is_outlier), nrow(res_B),
            100 * mean(res_B$is_outlier)))
mt_diagnose_clean_track(res_B)
```

Here Panel 1 has a single dominant mode, Panel 2 is flat, Panel 3 has
zero or near-zero bars, Panel 4 plateaus at iteration 1, and Panel 5
shows mostly run length 1 (or no flags at all). The diagnostic prints "no
concerns flagged" — the run was healthy.

## A multi-individual study

When you pass a multi-track result, panels 1–5 focus on the individual
with the highest flag rate by default, and panel 6 adds a per-individual
dotplot so you can see at a glance which animals to inspect next. Pass
`individual = "..."` to focus a specific track.

```{r multi-demo, eval = FALSE}
res_multi <- mt_clean_track(study, mass = my_mass_named, mode = "flying",
                              compact = TRUE, plot = FALSE, remove = FALSE)

## Default focus = highest-flag-rate individual
mt_diagnose_clean_track(res_multi)

## Or focus on a specific bird
mt_diagnose_clean_track(res_multi, individual = "Benjamin 7206 E0706")
```

## A decision tree for what to do next

| Diagnostic signature                              | Likely cause                          | Remedy                                                                                  |
|---|---|---|
| Multiple substantive modes (Panel 1)              | Bimodal behaviour (rest + flight)     | State-conditional analysis (see `vignette("OUTLIER_3_state_conditional", package = "move2utils")`); or filter to one state |
| Sustained band of elevated flag rate (Panel 2)    | Migration period being mistaken       | Filter the migration window or run it through state-conditional analysis                |
| Single-check bars dominate (Panel 3)              | One check noisy on this track; the combined decision did not promote it | Inspect those flags manually; consider raising that check's threshold                   |
| Iteration not plateauing (Panel 4)                | Self-reinforcing flagging             | Check Panel 1; supply a hard `v_max`; or use state-conditional analysis                 |
| Long-run tail without `error_class = "block"` (Panel 5) | Sustained behavioural state | Same as Panel 2 / 4                                                                     |
| Several individuals at >2 % (Panel 6, multi)      | Cohort heterogeneity                  | Triage list; run state-conditional or hard-cap path on the >5 % individuals             |

When several panels light up together with the same root cause — bimodal
distribution, a sustained migration band, non-converging iteration, a
heavy run-length tail — the answer is almost always state-conditional
analysis.

## The error_class taxonomy

Whichever consensus rule decides the flags, `mt_clean_track()` labels each
flagged fix with one `error_class`, so the diagnostic (and your own
inspection) reads the same vocabulary every time. The categories are:

- `consensus` — at least three of the four checks agreed.
- `geometric_spike` — the path-position and there-and-back checks agreed
  (the classic out-and-back spike).
- `state_anomaly` — a geometric check plus the speed cap.
- `kinematic_confluence` — a geometric check plus the unusual-movement
  check.
- `block` — part of a connected group of fixes cut off from the real
  trajectory and removed as a unit (the spoof / multi-hour-jump case).
- `physiological` — removed by a hard physiological speed cap (the
  `v_max` / `mass` + `mode` path).
- `state_transition_buffered` — a flag near a behavioural-state boundary
  that the buffer downgraded, so you can tell genuine errors from
  edge-of-state artefacts.

These labels are computed independently of the consensus rule, so
switching from the default `evidence_corroborated` to `class_aware`,
`strict`, `majority`, or any other rule changes *which* fixes are
flagged, never *how* a flagged fix is named.

## What the diagnostic does *not* do

It does not re-run the per-location checks. The path-position (bridge-η)
and joint-probability distributions — which would let you see whether a
score threshold landed in a real density valley or cut through a
continuum — are not part of the suite yet; a `recompute = TRUE` panel for
those is a planned extension.

It also assumes a single behavioural-state threshold was applied. Tracks
that you have already segmented by state and cleaned per segment will read
"healthy" on every panel, because each segment is unimodal.

It does not produce the `combined_evidence` column. That column is
returned only by the standalone `mt_flag_consensus()` (in its
`evidence_corroborated` or `weighted_evidence` mode); `mt_clean_track()`
returns the flag columns the diagnostic reads, not the evidence score.

## Further reading

- `vignette("OUTLIER_1_getting_started", package = "move2utils")` — the
  unified `mt_clean_track()` workflow and a brief tour of all four checks.
- `vignette("OUTLIER_3_state_conditional", package = "move2utils")` — when the
  diagnostic flags bimodal behaviour, the recipe for cleaning each
  behavioural state separately.
- `vignette("OUTLIER_4_outlier_bridge", package = "move2utils")` — the
  path-position check and the directional error-morphology classifier; for
  users who want fine-grained control over just one check.
- `vignette("OUTLIER_5_persistence_score", package = "move2utils")` —
  multi-scale annotation that scores how confidently each flag is an
  outlier; useful as a post-cleaning confidence filter.
- `vignette("OUTLIER_heterogeneous_error_regimes", package = "move2utils")` —
  outlier detection with heterogeneous error regimes: one sensor at a time.
- `vignette("OUTLIER_example_outlier_whitestork", package = "move2utils")` — a
  full narrated cleaning pipeline on a real high-frequency stork track.
- `vignette("OUTLIER_example_leo_migration", package = "move2utils")` — outlier
  detection on irregular, large-scale satellite data.
</content>
</invoke>
