---
title: "Producing one UD per behavioural state (day/night, season, etc.)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Producing one UD per behavioural state (day/night, season, etc.)}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 5,
  message = TRUE
)
```

Sometimes you don't want one UD per animal — you want one UD per
**context**: this bird's daytime UD vs. its nighttime UD, this
animal's summer range vs. its winter range, this seal's foraging
UD vs. its haul-out UD. The pattern: you have a category that
varies *within* one track, and you want a separate UD for each
category, stacked together on a common grid so they're directly
comparable.

This vignette shows two ways to do that with `mt_dbbmm_ud()` and
`mt_dbgb_ud()`. The first uses the existing multi-track dispatch
(by attaching the context to the track ID); the second uses
`split()` + `lapply` when each context needs different settings.

(If you used the legacy `move` package's `MoveBurst` class, this
is the `move2` equivalent. `move2` deliberately doesn't reproduce
`MoveBurst` — context belongs on the tibble as a column, and the
per-context workflow is expressible directly with `dplyr` and the
package's normal multi-track dispatch. The result is the same kind
of stacked UD output you'd get from `DBBMMBurstStack`.)

## Two idioms

### A. Composite `track_id`

Make the context part of the track identifier. The existing
multi-track dispatch of `mt_dbbmm_ud()` / `mt_dbgb_ud()` then does
exactly what `MoveBurst` used to do: one UD per composite label,
stacked on a common grid.

### B. `split()` and iterate

Keep the original `track_id`. Split the object into a list of
sub-tracks, one per context, and `lapply` the UD function. Needed
when contexts require their own grids or settings.

A is the default; B is the fallback.

## Setting the scene

```{r load, message = FALSE}
library(move2)
library(dplyr)
library(sf)
library(terra)
library(move2utils)

fishers <- mt_read(mt_example())
fishers <- fishers[!st_is_empty(fishers), ]

## three individuals with enough data for a reasonable per-burst UD
ids <- c("F1", "F2", "M1")
focal <- filter_track_data(fishers, .track_id = ids)
focal <- focal |> dplyr::group_by(mt_track_id(focal)) |> slice(1:500) |> ungroup()
## any CRS works: the UD functions auto-project and return in the input
## CRS, so we keep this lon/lat track as is
```

## Attaching a per-location context column

The simplest contexts are per-location (length `n`): derived from the
timestamp (hour-of-day, season), from an environmental annotation
(habitat class), or from a downstream classifier (HMM state). Attach
them as columns.

```{r day-night}
focal <- focal |>
  dplyr::mutate(
    hour       = as.integer(format(mt_time(focal), "%H")),
    day_night  = factor(
      dplyr::if_else(hour >= 6 & hour < 18, "day", "night"),
      levels = c("day", "night")
    )
  )
table(focal$day_night, mt_track_id(focal))
```

For per-**segment** contexts (length `n-1`), align them to the **starting**
location of each segment and pad the final location with `NA` or with
the last segment's label — whichever matches your intent.

## Idiom A — composite `track_id`

Build a composite label and set it as the track identifier. Existing
multi-track dispatch handles the rest.

```{r composite}
focal_burst <- focal |>
  dplyr::mutate(
    burst_id = paste(mt_track_id(focal), day_night, sep = "__")
  )
focal_burst <- mt_set_track_id(focal_burst, "burst_id")

table(mt_track_id(focal_burst))
```

### A dBBMM UD per `(individual, day/night)` pair

`mt_dbbmm_variance()` accepts a multi-track `move2` object and
returns a named list of variance objects. `mt_dbbmm_ud()` on that
list rasters each UD on a common grid computed from the combined
extent, so the layers are directly comparable.

```{r dbbmm-per-burst}
var_list <- mt_dbbmm_variance(
  focal_burst,
  location_error = 25,
  window_size    = 31,
  margin         = 11
)
length(var_list)
names(var_list)
```

```{r ud-per-burst, fig.width = 9, fig.height = 6, fig.cap = "Day / night UDs for F1, F2, M1. Each panel is an Cumulative volume UD on a common grid, so the layers are directly comparable."}
stk <- mt_dbbmm_ud(var_list, location_error = 25, raster = 100,
                    ext = 0.85, verbose = FALSE)
names(stk)
vud <- ud_volume(stk)
terra::plot(vud)
```

Each layer of `stk` is an ordinary UD with cells summing to 1. To
compare "F1 day" against "F1 night" you index the stack with its
name: `stk[["F1__night"]]`.

### Caveats

- **Filter out short bursts first.** `mt_dbbmm_variance()` will error
  or warn on bursts that contain fewer locations than `window_size`.
  Compute `table(mt_track_id(focal_burst))` and drop sparse contexts
  before fitting.

- **Do not burst across long gaps.** If `day_night` flips in the
  middle of an overnight gap, the burst boundary artificially closes
  the UD at the two gap endpoints. Mask long-gap segments with
  `mt_mask_segments()` *after* fitting variance, or skip the
  transition points before labelling. See
  `vignette("UD_gap_aware_ud", package = "move2utils")` for the full
  workflow.

- **Common-grid discipline.** If you compare bursts visually or
  metrically (e.g. overlap), the multi-track dispatch already puts
  all layers on the same grid. If you mix outputs from separate
  dispatches, pass a pre-made `SpatRaster` to each call.

## Idiom B — `split()` and `lapply()`

When each burst needs its own grid, resolution, or parameters, split
the object explicitly and iterate.

```{r split-iterate, eval = FALSE}
bursts <- split(focal, interaction(mt_track_id(focal),
                                       focal$day_night,
                                       drop = TRUE))
bursts <- bursts[sapply(bursts, nrow) >= 60]  # drop sparse bursts

var_each <- lapply(bursts, mt_dbbmm_variance,
                    location_error = 25,
                    window_size    = 31,
                    margin         = 11)

## compute each UD on a common grid so they remain stackable
all_xy <- do.call(rbind, lapply(var_each, function(v) {
  data.frame(x = v$track_data$x, y = v$track_data$y)
}))
pts <- sf::st_as_sf(all_xy, coords = c("x", "y"),
                     crs = var_each[[1]]$track_data$crs)
grid <- move2utils:::.make_raster(pts, cell_size = 100, ext = 0.85)

ud_each <- lapply(var_each, mt_dbbmm_ud,
                   location_error = 25, raster = grid, verbose = FALSE)
stk_b <- terra::rast(ud_each)
names(stk_b) <- names(bursts)
```

This pattern is also what you want when the context implies genuinely
different `location_error` or `window_size` per burst — for example a
tag that switched duty cycle mid-study.

## dBGB bursts for directionality contrasts

Everything above works unchanged for `mt_dbgb_variance()` and
`mt_dbgb_ud()`. dBGB is particularly informative for per-burst
contrasts because it separates directed from random space use.
Consider a commuting animal where day-time movement is foraging
(parallel ≈ perpendicular variance) and night-time is commuting
(parallel ≫ perpendicular):

```{r dbgb-per-burst, eval = FALSE}
var_list_d <- mt_dbgb_variance(focal_burst, location_error = 25,
                                 window_size = 31, margin = 11)
stk_d <- mt_dbgb_ud(var_list_d, location_error = 25,
                     raster = 100, ext = 0.85)
```

Contrast the parallel/perpendicular ratios across bursts via
`mt_motion_variance()` on each list element — the ratio is the
diagnostic signal that dBGB adds on top of dBBMM.

## When a burst should *not* be a new track

Some contexts are not really bursts at all — they are filters. For
instance, to compute a UD only for locations in a specific habitat,
the right operation is `filter()`, not `burst()`. If the animal never
returns to the filtered class, the UD is defined. If it does, you
have *three* options and should choose deliberately:

1. Compute the UD on the filtered subset only (you get habitat-specific
   use).
2. Compute the UD on the whole track and then *mask* cells outside the
   habitat of interest.
3. Compute the UD on the whole track, unmasked, and report the
   fraction of the UD falling in the habitat.

These answer different questions; the difference is not a detail.

## Further reading

- `vignette("UD_dbbmm_ud", package = "move2utils")` — the Dynamic Brownian 
bridge UDs (dBBMM).
- `vignette("UD_dbgb_ud", package = "move2utils")` — the Directional UDs with 
the dynamic bivariate Gaussian bridge (dBGB).
- `vignette("UD_gap_aware_ud", package = "move2utils")` — excluding long-gap 
segments from dBBMM/dBGB
- `vignette("UD_ud_comparison", package = "move2utils")` — comparing UDs, stacking 
on a common grid, computing an overlap metric, 
and computing Earth-mover's distance (EMD).
