---
title: "Directional UDs with the dynamic bivariate Gaussian bridge (dBGB)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Directional UDs with the dynamic bivariate Gaussian bridge (dBGB)}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 5,
  message = TRUE
)
```

When an animal travels in a committed direction — a migration leg,
a daily commute between roost and foraging site — you intuitively
expect its utilisation distribution to be elongated *along* that
direction, not a round blob. The standard dynamic Brownian bridge
(dBBMM) doesn't capture this: it treats movement between two fixes
as wandering equally in all directions, so it spreads the UD into
a circle around each step. For a directed traveller, that
over-spreads sideways and under-represents the actual route.

The **dynamic bivariate Gaussian bridge (dBGB)** fixes this. It
splits the bridge between two fixes into two parts: one along the
direction of travel and one perpendicular to it. Each part has its
own variance. When the animal is moving fast and straight, the
along-track variance is large (it could be anywhere on the line)
but the across-track variance is small (it's not off to one side).
The resulting UD hugs the travel axis instead of spreading into a
disc.

`mt_dbgb_variance()` and `mt_dbgb_ud()` are the directional
counterparts to `mt_dbbmm_variance()` and `mt_dbbmm_ud()`. They
share the same two-step workflow (fit the variance, then rasterise
the UD) and return UDs on the same kind of `terra` grid, so you
can swap them in without restructuring downstream code. If you
used `move::brownian.bridge.dyn()` before, the dBGB is the
directional generalisation: when the parallel and perpendicular
variances are equal, dBGB collapses back to dBBMM.

## When to reach for dBGB rather than dBBMM

| Movement mode | Reach for |
|---|---|
| Foraging, resting, random area use | `mt_dbbmm_*` (isotropy is a fine prior) |
| Commuting, migration legs, directed travel | `mt_dbgb_*` |
| Mixed (the usual case) | Either. dBGB generalises dBBMM: when parallel and perpendicular variances are equal, the two agree. It is safe to default to dBGB if you suspect directionality matters. |

The cost is modest — roughly twice the compute of dBBMM because two
variances have to be estimated per window — and the benefit is a UD
that respects the direction of travel.

## Load a track

```{r load, message = FALSE}
library(move2)
library(sf)
library(dplyr)
library(terra)
library(move2utils)

fishers <- mt_read(mt_example())
fishers <- fishers[!st_is_empty(fishers), ]

## Leroy again, first 500 fixes for a fast build.  Like dBBMM, the dBGB
## functions accept any CRS and return the UD in it, so we pass this
## lon/lat track straight in (project first with mt_aeqd_crs() if you
## prefer a metric grid).
leroy <- filter_track_data(fishers, .track_id = "M1")
leroy <- slice(leroy,1:500)
```

## Fit a dBGB

The signature mirrors dBBMM. `mt_dbgb_variance()` returns an S3 object
holding the parallel and perpendicular standard deviations; `mt_motion_variance()`
extracts them as a `data.frame` with columns `para` and `orth`.

```{r variance}
var_dbgb <- mt_dbgb_variance(
  leroy,
  location_error = 20,
  window_size    = 31,
  margin         = 11
)

## per-location directional variances
v_df <- mt_motion_variance(var_dbgb)
summary(v_df)
```

A quick comparison of the two dimensions over time tells you whether
the animal is directional. The ratio `para / orth > 1` indicates
elongation along the direction of travel.

```{r variance-plot}
time_days <- as.numeric(mt_time(leroy) - mt_time(leroy)[1], units = "days")
plot(time_days, v_df$para, type = "l", col = "steelblue", lwd = 1.5,
     ylim = c(0, max(v_df, na.rm = TRUE)),
     xlab = "time (days since start)", ylab = "variance (m^2)",
     main = "parallel vs perpendicular variance")
lines(time_days, v_df$orth, col = "firebrick", lwd = 1.5)
legend("topright", c("parallel", "perpendicular"),
       col = c("steelblue", "firebrick"), lwd = 1.5, bty = "n")
```

## Rasterise the UD

`mt_dbgb_ud()` accepts the variance object (or a `move2` track
directly, in which case it estimates variance internally) and
produces a `SpatRaster` whose cells sum to 1.

```{r ud}
ud_dbgb <- mt_dbgb_ud(var_dbgb, location_error = 20,
                       raster = 100, ext = 0.85)

terra::plot(ud_dbgb, main = "dBGB utilisation distribution")
```

If you see `The raster does not extent far enough ...`, increase `ext`
(padding around the auto-generated raster) or pass a pre-made
`SpatRaster` that is large enough. The dBGB kernel needs more margin
than dBBMM when the parallel variance is high, because the bridge
extends further along the direction of travel.


## Cumulative-volume UD and isopleths

`ud_volume()` converts the probability-density UD into cumulative-
volume space and works on any UD `SpatRaster`, dBBMM or dBGB (see
`vignette("UD_dbbmm_ud", package = "move2utils")` for more details).

```{r volume}
vud_dbgb <- ud_volume(ud_dbgb)
terra::plot(vud_dbgb, main = "dBGB cumulative volume")
terra::contour(vud_dbgb, levels = c(0.5, 0.95), add = TRUE,
               lty = c(2, 1), lwd = c(0.5, 0.5))
```

## Side-by-side: dBBMM vs dBGB on the same track

A two-panel comparison reveals where the two models differ. Same
track, same grid, same `location_error`:

```{r compare, fig.height = 5}
var_dbbmm <- mt_dbbmm_variance(
  leroy, location_error = 20,
  window_size = 31, margin = 11
)
ud_dbbmm <- mt_dbbmm_ud(var_dbbmm, location_error = 20, raster = 100,
                         ext = 0.85, verbose = FALSE)

## put both UDs on the same extent so the two panels are comparable
common_ext <- terra::union(terra::ext(ud_dbbmm), terra::ext(ud_dbgb))
ud_dbbmm <- terra::extend(ud_dbbmm, common_ext)
ud_dbgb  <- terra::extend(ud_dbgb,  common_ext)

## plotting the cumulative-volume UD for better visualization
vud_dbbmm <- ud_volume(ud_dbbmm)
vud_dbgb <- ud_volume(ud_dbgb)
par(mfrow = c(1, 2))
terra::plot(vud_dbbmm, main = "dBBMM (isotropic)")
terra::plot(vud_dbgb,  main = "dBGB (directional)")
```

On a fisher, which foraged extensively over this window, the two
UDs will look similar. On a commuting or migrating animal, the
dBGB UD narrows visibly along the travel legs. Try the Leo turkey
vulture example (`vignette("OUTLIER_example_leo_migration", package = "move2utils")`)
for a clearer case.

## Gap handling

dBGB exposes a segment-level interest flag (`seg_interest`) rather than
a location-level `interest` flag. The semantics are the same — a
segment marked `FALSE` is skipped during UD integration — but you
index it one shorter than the track length (n-1 segments).

```{r gap, eval = FALSE}
lag_min <- as.numeric(mt_time_lags(leroy, units = "min"))
var_dbgb$seg_interest[lag_min > 300] <- FALSE

ud_masked <- mt_dbgb_ud(var_dbgb, location_error = 20, raster = 100,
                         ext = 0.85, verbose = FALSE)
```

## Further reading

- `vignette("UD_dbbmm_ud", package = "move2utils")` — the Dynamic Brownian 
bridge UDs (dBBMM).
- `vignette("UD_bursted_uds", package = "move2utils")` — producing dBBMM and 
dBGB UDs per behavioural or temporal segment per individual.
- `vignette("UD_gap_aware_ud", package = "move2utils")` — excluding long-gap 
segments from dBBMM/dBGB
- `vignette("UD_ud_comparison", package = "move2utils")` — comparing UDs, stacking 
on a common grid, computing an overlap metric, 
and computing Earth-mover's distance (EMD).
- Kranstauber (2019). Modelling animal movement as Brownian bridges
  with covariates. *Movement Ecology* 7:22.
