---
title: "Dynamic Brownian bridge UDs (dBBMM)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dynamic Brownian bridge UDs (dBBMM)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 5,
  message = TRUE
)
```

Here we present a workflow for the calculation of dynamic Brownian bridge 
movement models (dBBMM) and UDs. We have updated the dBBMM related functions 
from the `move` package to use `move2`, `sf` and `terra`, and in the process 
also improved the algorithm which is now much faster.
If you have migrated from `move`, the translation is:

| legacy `move` | `move2utils` |
|---|---|
| `spTransform(x, center = TRUE)` | not needed — any CRS works (optionally `sf::st_transform(x, move2::mt_aeqd_crs(x))` for a metric grid) |
| `brownian.motion.variance.dyn()` | `mt_dbbmm_variance()` |
| `brownian.bridge.dyn()` | `mt_dbbmm_ud()` (or the two-step variance → UD) |
| `getMotionVariance()` | `mt_motion_variance()` |
| `getVolumeUD()` | `ud_volume()` (or the inline recipe shown below) |

There are two main functions in the workflow: `mt_dbbmm_variance()`
estimates a dynamic Brownian-motion variance per location, and
`mt_dbbmm_ud()` rasterises the bridge-weighted UD on a common grid.

## Load a track

```{r load, message = FALSE}
library(move2)
library(sf)
library(dplyr)
library(terra)
library(move2utils)

## get move2 example data set: fishers GPS data
fishers <- mt_read(mt_example())
fishers <- fishers[!st_is_empty(fishers), ]

## Leroy (M1); keep the first few hundred
## locations so the vignette builds quickly
leroy <- filter_track_data(fishers, .track_id = "M1")
leroy <- slice(leroy,1:500)

## mt_dbbmm_variance() and mt_dbbmm_ud() accept a track in any CRS and
## return the UD in that same CRS, so we pass this lon/lat track straight
## in. (The bridge math runs internally in a local equal-area projection;
## if you would rather work on a metric grid yourself, project first with
## st_transform(leroy, mt_aeqd_crs(leroy)) and the UD comes back in it.)
nrow(leroy)
```

## Fit a dBBMM

The two-step workflow separates variance estimation from UD
rasterisation so that you can reuse the variance object when tuning
the grid. Pass `location_error` in map units (metres here); window
size and margin control the sliding estimator (see
`?mt_dbbmm_variance`). Note that in most cases different values of window size 
and the margin do not have a great effect on the results.

```{r variance}
var_obj <- mt_dbbmm_variance(
  leroy,
  location_error = 20,
  window_size    = 31,
  margin         = 11
)

## a numeric vector of per-location variances (NA at track margins)
head(mt_motion_variance(var_obj))
```

Rasterise the UD. There are two options, either set the cell size of the raster, 
e.g. `raster = 100` requests a 100 m cell size; or pass a
pre-made `SpatRaster` when you need UDs from different tracks on a
shared grid.

```{r ud}
ud <- mt_dbbmm_ud(var_obj, location_error = 20, raster = 100)

terra::plot(ud, main = "dBBMM utilisation distribution")
```

The single-step shortcut `mt_dbbmm_ud(leroy, location_error = 20)`
fits the variance internally — convenient when you do not need the
variance object itself.

The sum of all pixels of the dBBMM UD raster sum up to 1. Plotting the result 
provides limited visual information, as the values correspond to the probability 
the animal was present in a given pixel during the observed period. These results 
can be used to estimate probability of the animal being present in a certain area 
during the tracking period (see 'Cumulative-volume UD' section), or to calculate 
the probability of the animal being present in a certain environment during the 
tracking period by overlaying the resulting raster over a categorical raster 
(e.g. of land cover) and summing up the dBBMM pixel values that fall within 
each category.

## Cumulative-volume UD and isopleths

`ud_volume()` converts a UD into cumulative-probability space, which represents 
the minimum area in which an animal has some specified probability of being 
located. (This is the `move2` analogue of `move::getVolumeUD()`.)

```{r volume}
vud <- ud_volume(ud)
terra::plot(vud, main = "Cumulative volume UD")
terra::contour(vud, levels = c(0.5, 0.95), add = TRUE,
               lty = c(2, 1), lwd = c(0.5, 0.5))
```

Solid 50 % and 95 % masks are simple logical operations on the
volume UD:

```{r isopleths}
ud50 <- vud <= 0.50
ud95 <- vud <= 0.95

par(mfrow = c(1, 2))
terra::plot(ud95, main = "95% area", col = c("grey90", "steelblue"))
terra::plot(ud50, main = "50% area", col = c("grey90", "firebrick"))
```

If you need the isopleth as a vector (line or polygon), `terra`'s own
`as.contour()` returns a `SpatVector` that coerces cleanly to `sf`:

```{r contour-sf, eval = FALSE}
iso95 <- terra::as.contour(vud, levels = 0.95) |>
  sf::st_as_sf()
```

## Handling gaps and the time-step pitfall

### Choose a sensible `time_step`

`mt_dbbmm_ud()` integrates the Brownian bridge along each segment. The
step defaults to 1/15 of the smallest time lag in the track, which
keeps the quadrature stable for short segments. Inspect your time
lags before fitting so you know what you're working with:

```{r lags}
lags <- as.numeric(mt_time_lags(leroy, units = "min"))
summary(lags)
```

If you have a track with a median lag much larger than the minimum,
and very few short lags, you can override `time_step = median(lags)/15`
to speed things up; but keep the *smallest* lag governing the step
when short segments carry real information about rapid movement.

### Do not bridge across long gaps

A missing satellite pass leaves a long time lag between otherwise
adjacent fixes. Integrating the bridge across that gap smears
probability across regions the animal never visited. The fix is to
mark those locations as "not of interest" in the variance object
before rasterising; the UD integration will skip them.

We remove a block of fixes to open up one long (~26 h) gap. Leroy's
first 500 fixes also already contain several natural overnight gaps, so
masking every segment longer than 300 minutes cleans all of them at
once — the injected gap included. Compare the UD with and without the
mask:

```{r gap-demo}
## remove a block of fixes to open a long (~26 h) gap
keep <- setdiff(seq_len(nrow(leroy)), 200:260)
leroy_gap <- leroy[keep, ]

var_gap <- mt_dbbmm_variance(
  leroy_gap, location_error = 20,
  window_size = 31, margin = 11
)

## the UD with no gap masking — probability bleeds across the gap
ud_bleed <- mt_dbbmm_ud(var_gap, location_error = 20, raster = 100,
                         verbose = FALSE)

## mask out locations whose following segment exceeds 300 minutes
lag_gap <- as.numeric(mt_time_lags(leroy_gap, units = "min"))
var_gap$interest[which(lag_gap > 300)] <- FALSE

ud_masked <- mt_dbbmm_ud(var_gap, location_error = 20, raster = 100,
                          verbose = FALSE)

## for better visualization plotting the cumulative volume UD
par(mfrow = c(1, 2))
terra::plot(ud_volume(ud_bleed),  main = "unmasked: bleeds across gap")
terra::plot(ud_volume(ud_masked), main = "masked: clean")
```

The difference is usually visible as a low-density ribbon connecting
the two sub-tracks; masking removes it.

## Multi-individual UDs on a common grid

Both `mt_dbbmm_variance()` and `mt_dbbmm_ud()` dispatch on
multi-track `move2` objects: they return a named list of variance
objects and a multi-layer `SpatRaster` respectively, all on a grid
computed from the combined extent so that layers are directly
comparable.

```{r multi, eval = FALSE}
three <- fishers[mt_track_id(fishers) %in% c("F1", "F2", "M1"), ]

## multi-track input is auto-projected to one shared local grid and the
## stack is returned in the input CRS (here lon/lat)
stk <- mt_dbbmm_ud(three, location_error = 25,
                    window_size = 31, margin = 11)
names(stk)
terra::plot(stk)
```

## Persistence

The package does not ship its own IO helpers — the `sf` and `terra`
ecosystems already provide every format the legacy `move` vignette
covered (shapefile, GeoTIFF, GeoPackage, KML, CSV).

```{r io, eval = FALSE}
saveRDS(var_obj, "leroy_dbbmm_variance.rds")
terra::writeRaster(ud, "leroy_ud.tif", overwrite = TRUE)
sf::st_write(leroy, "leroy.gpkg", delete_layer = TRUE)
```

## Further reading

- `vignette("UD_dbgb_ud", package = "move2utils")` — the Directional UDs with 
the dynamic bivariate Gaussian bridge (dBGB).
- `vignette("UD_bursted_uds", package = "move2utils")` — producing dBBMM and 
dBGB UDs per behavioural or temporal segment per individual.
- `vignette("UD_gap_aware_ud", package = "move2utils")` — excluding long-gap 
segments from dBBMM/dBGB
- `vignette("UD_ud_comparison", package = "move2utils")` — comparing UDs, stacking 
on a common grid, computing an overlap metric, 
and computing Earth-mover's distance (EMD).
- Kranstauber, B., Kays, R., LaPoint, S. D., Wikelski, M., & Safi, K.
  (2012). A dynamic Brownian bridge movement model to estimate
  utilization distributions for heterogeneous animal movement.
  *Journal of Animal Ecology*, 81(4), 738–746.
