---
title: "Gap-aware UDs: excluding long-gap segments from dBBMM/dBGB"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gap-aware UDs: excluding long-gap segments from dBBMM/dBGB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 5,
  message = TRUE
)
```

When you fit a dBBMM (or a dBGB) on a track with long inter-fix
gaps, you get a problem. The bridge between two fixes that are
12 hours apart is much *wider* than the bridge between two fixes
15 minutes apart — that's by design, it reflects the fact that
the animal had more time to wander between them. But when the
animal was actually denning, sleeping in a burrow, or out of GPS
view, *it didn't wander*. The wide bridge spreads the UD across a
region the animal never visited. The smear is a pure interpolation
artefact driven by elapsed time, not by evidence of movement.

This vignette shows two ways to handle that:

- **Mask the long-gap segments** (the simpler option, when you
  just want one clean UD). You compute the variance object as
  normal, then tell the UD step "ignore the bridges across these
  gap segments" via `mt_mask_segments()`. The UD comes out
  without the smear.
- **Burst around the gaps** (the right option when the separate
  runs between gaps are themselves of interest, or when you want
  to compare per-run dBGB direction). You make each run a separate
  composite `track_id` and let the multi-track dispatch produce
  one UD per run, following the idiom from
  `vignette("UD_bursted_uds", package = "move2utils")`.

(If you used the legacy `move` package, this is the modern
equivalent of setting `interested = FALSE` on the variance object.
The mask-based workflow does the same thing under the hood —
writing to the per-segment include-mask — just behind a single
helper that works identically on dBBMM and dBGB objects.)

The running example uses the fisher GPS data bundled with `move2`
and a six-hour gap threshold, which roughly matches a fisher's
denning duration and trips on 19 of Leroy's 918 segments.

## Setting up

```{r load, message = FALSE}
library(move2)
library(dplyr)
library(sf)
library(terra)
library(units)
library(move2utils)

fishers <- mt_read(mt_example())
fishers <- fishers[!st_is_empty(fishers), ]

## focus on Leroy (M1).  The dBBMM/dBGB functions auto-project internally
## and return the UD in the input CRS, so we pass this lon/lat track in as
## is (project with mt_aeqd_crs() first if you prefer a metric grid).
leroy <- filter_track_data(fishers, .track_id = "M1")

nrow(leroy)
```

## Inspecting the gap distribution

```{r gap-distribution}
lag_hours <- as.numeric(set_units(mt_time_lags(leroy), "hour"))
summary(lag_hours)

threshold_h <- 6
long_gap    <- which(lag_hours > threshold_h)
length(long_gap)
```

The last element of `mt_time_lags()` is `NA` (no segment after the
final location). Segments longer than the threshold are the candidates
for exclusion: their endpoints are widely separated with no
intermediate movement information, so the bridge is a pure
interpolation across territory that the animal may or may not have
crossed.

## Approach A — mask long-gap segments

Fit the variance on the full track, so the surrounding dynamics are
estimated with their normal window context, then pass the long-gap
segment indices to `mt_mask_segments()`. The UD routine respects the
mask and simply skips those bridges.

```{r approach-A}
var_full <- mt_dbbmm_variance(
  leroy,
  location_error = 25,
  window_size    = 31,
  margin         = 11
)

var_masked <- mt_mask_segments(var_full, long_gap)
```

```{r ud-A, fig.width = 11, fig.height = 4, fig.cap = "Left: naive UD fits a bridge across every segment, including the 19 gaps longer than six hours, spreading probability over areas the animal never visited. Middle: long-gap bridges masked out with mt_mask_segments(); probability re-concentrates onto the true track. Right: difference (naive − masked) isolates the smear artefact — positive cells are the fake density the gap bridges deposited."}
ud_smeared <- mt_dbbmm_ud(var_full,   location_error = 25,
                            raster = 100, ext = 0.5, verbose = FALSE)
ud_clean   <- mt_dbbmm_ud(var_masked, location_error = 25,
                            raster = 100, ext = 0.5, verbose = FALSE)
ud_diff    <- ud_smeared - ud_clean

common_range <- range(c(terra::values(ud_smeared),
                         terra::values(ud_clean)), na.rm = TRUE)

par(mfrow = c(1, 3), mar = c(2, 2, 3, 4))
terra::plot(ud_smeared, range = common_range,
             main = "Naive UD — fans out over gaps")
terra::plot(ud_clean, range = common_range,
             main = "Gap-masked UD")
terra::plot(ud_diff, main = "Difference (naive − masked)")
```

The difference panel is the key diagnostic: positive cells mark the
probability mass that the long-gap bridges deposited in non-visited
space — pure interpolation artefact that the mask removes. The same
code, unchanged, applies to a dBGB variance object:

```{r approach-A-dbgb, eval = FALSE}
var_full_d   <- mt_dbgb_variance(leroy, location_error = 25,
                                   window_size = 31, margin = 11)
var_masked_d <- mt_mask_segments(var_full_d, long_gap)
ud_clean_d   <- mt_dbgb_ud(var_masked_d, location_error = 25,
                             raster = 100, ext = 0.5)
```

`mt_mask_segments()` dispatches on the object class and writes to the
appropriate include-mask. Multi-track variance lists are supported too:
passing the same `long_gap` vector broadcasts it across all elements,
or a named list of segment vectors applies a per-track mask.

## Approach B — burst around each gap

When the individual runs between gaps are the unit of analysis — e.g.
contrasting pre- and post-denning space use, treating each nightly
foray as its own UD, or pairing with a dBGB run-level directional
summary — wrap each run in its own composite `track_id`. Denning
segments then occupy their own (two-point) bursts that are dropped by
the `window_size` filter.

```{r burstify}
## burst_ix increments each time a long gap is crossed;
## a location is assigned to the burst that contains its arrival.
is_long_gap_in <- c(FALSE, head(lag_hours, -1) > threshold_h)
is_long_gap_in[is.na(is_long_gap_in)] <- FALSE
burst_ix       <- cumsum(is_long_gap_in)

leroy_b <- leroy |>
  mutate(
    burst_id = sprintf("%s__run%02d", mt_track_id(leroy), burst_ix)
  )
leroy_b <- mt_set_track_id(leroy_b, "burst_id")

## drop runs too short to fit the window;
## with window_size = 31 the minimum is 31, but 40 leaves a small margin.
min_run <- 40
short   <- names(which(table(mt_track_id(leroy_b)) < min_run))
leroy_b <- leroy_b[!mt_track_id(leroy_b) %in% short, ]

table(mt_track_id(leroy_b))
```

```{r variance-B, eval = FALSE}
var_list <- mt_dbbmm_variance(
  leroy_b,
  location_error = 25,
  window_size    = 31,
  margin         = 11
)

## per-run UDs rastered on a common grid
stk <- mt_dbbmm_ud(var_list, location_error = 25, raster = 100, ext = 0.5)
names(stk)
terra::plot(stk)
```

To collapse the per-run stack into a single "active-space" UD, weight
each run by its duration and sum:

```{r combine-B, eval = FALSE}
durations <- vapply(var_list, function(v) {
  diff(range(v$track_data$time_mins))    # minutes
}, numeric(1))
w <- durations / sum(durations)

ud_active <- sum(stk * w)
ud_active <- ud_active / sum(terra::values(ud_active), na.rm = TRUE)
terra::plot(ud_active, main = "Duration-weighted active-space UD")
```

Use a location-count weighting (number of locations per run) instead
of duration when the time a run spans is confounded by sampling-rate
changes.

The bursting idiom works identically for `mt_dbgb_variance()` /
`mt_dbgb_ud()`. dBGB is particularly informative here: per-run
parallel/perpendicular ratios from `mt_motion_variance()` distinguish
commuting-style runs (parallel ≫ perpendicular) from foraging-style
runs (parallel ≈ perpendicular), which is typically the reason one
wants runs separated rather than collapsed in the first place.

## Choosing between approaches

- **Approach A (mask).** One holistic UD with bad segments excised.
  Cheapest: a single variance fit and a single UD rasterisation. The
  sliding window still spans the gap during variance estimation, so
  neighbouring per-location variance estimates remain mildly affected
  by the gap's coordinates — a minor second-order effect, worth
  flagging when gaps are frequent.
- **Approach B (burst).** One independent variance fit per run, and
  therefore no cross-gap leakage in the variance estimates. More
  expensive, and requires deciding how to combine (or whether to
  combine) the per-run UDs. Worth it when runs themselves are the
  unit, or whenever gaps dominate the track enough that Approach A's
  leakage becomes non-negligible.

For typical GPS tracks with a handful of overnight gaps, Approach A is
the right first move. For tracks with many gaps, frequent denning, or
a clear behavioural rhythm mapped onto gap/no-gap, Approach B is the
more conservative default.

## Further reading

- `vignette("UD_dbbmm_ud", package = "move2utils")` — the Dynamic Brownian 
bridge UDs (dBBMM).
- `vignette("UD_dbgb_ud", package = "move2utils")` — the Directional UDs with 
the dynamic bivariate Gaussian bridge (dBGB).
- `vignette("UD_bursted_uds", package = "move2utils")` — producing dBBMM and 
dBGB UDs per behavioural or temporal segment per individual.
- `vignette("UD_ud_comparison", package = "move2utils")` — comparing UDs, stacking 
on a common grid, computing an overlap metric, 
and computing Earth-mover's distance (EMD).
