---
title: "Comparing utilisation distributions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing utilisation distributions}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 5,
  message = TRUE
)
```

Once you have several UDs — for different individuals, different
seasons, or different behavioural contexts (see
`vignette("UD_bursted_uds", package = "move2utils")`) — the next
question is usually how similar they are. This vignette covers the
three comparison operations: stacking on a common grid, computing an 
overlap metric, and computing Earth-mover's distance (EMD).

## Stacking on a common grid

Every pairwise UD metric depends on the two UDs having identical
extents and cell geometry. `mt_dbbmm_ud()` and `mt_dbgb_ud()` enforce
this automatically on multi-track input. The only situation where
you need to intervene manually is when the UDs come from separate
dispatches (e.g. two datasets loaded at different times).

This is how you can get a stack with each layer renormalised to sum to 1:

```{r udstack-idiom, eval = FALSE}
## Given a list of UDs (or a multi-layer SpatRaster) on a common grid:
uds_list <- list(day = ud_day, night = ud_night)                # or split

stk <- terra::rast(uds_list)                                     # stack
stk <- stk / terra::global(stk, "sum", na.rm = TRUE)[, 1]        # renormalise
names(stk) <- names(uds_list)
```

The `global(..., "sum")` divisor is a per-layer scalar, so the
division broadcasts correctly and every output layer sums to 1 . If you
built the UDs with `mt_dbbmm_ud()` or `mt_dbgb_ud()`'s multi-track dispatch, or
with the composite-`track_id` idiom from
`vignette("UD_bursted_uds", package = "move2utils")`, the output is
already on a common grid and already per-layer normalised — you do
not need the renormalisation step.

```{r load, message = FALSE}
library(move2)
library(dplyr)
library(sf)
library(terra)
library(move2utils)

fishers <- mt_read(mt_example())
fishers <- fishers[!st_is_empty(fishers), ]

two <- filter_track_data(fishers, .track_id = c("F3", "M3"))
## keep each individual to its first 500 fixes for a fast build
two <- two |>
  dplyr::group_by(mt_track_id(two)) |>
  dplyr::slice_head(n = 500) |>
  dplyr::ungroup()
## The UD functions accept any CRS, but here we deliberately project to a
## metric grid: the EMD default ground metric is Euclidean (and overlap
## metrics compare cells), which only makes sense on equal-area cells.
## On a lon/lat grid you would pass gc = TRUE to emd() instead (see below).
two_p <- st_transform(two, mt_aeqd_crs(two))

stk <- mt_dbbmm_ud(
  two_p,
  location_error = 25,
  window_size    = 31,
  margin         = 11,
  raster         = 100,
  ext            = 1.25
)
names(stk)

```

`stk` and `vud` are both a two-layer `SpatRaster`; the layers share grid and CRS.

## Volume UD 
Most comparison operations work more cleanly on cumulative-volume
UDs than on raw probability densities. The transform orders cells by
probability density and replaces each value by the cumulative sum —
so cell *v* = 0.5 means "this cell is inside the 50 %-volume
contour, together with all denser cells". `ud_volume()` applies
the transform; it accepts a multi-layer `SpatRaster` and acts
layer-wise.

```{r vol-ud}
vstk <- ud_volume(stk)
names(vstk)
terra::plot(vstk)
```

## Overlap metrics

### Core-area overlap (Jaccard)

The overlap of the two *p*-volume contours — shared area divided by
combined area (a symmetric Jaccard index) — is a common "core-area
overlap" summary. At `p = 0.5` it compares the two 50 % cores.

```{r vol-intersect}
in_core <- function(vud_a, vud_b, p = 0.5) {
  mask_a <- vud_a <= p
  mask_b <- vud_b <= p
  sum(terra::values(mask_a & mask_b), na.rm = TRUE) /
    sum(terra::values(mask_a | mask_b), na.rm = TRUE)
}

in_core(vstk[[1]], vstk[[2]], p = 0.50)   # 50% core overlap
in_core(vstk[[1]], vstk[[2]], p = 0.95)   # 95% use overlap
```

### Bhattacharyya affinity

A density-level overlap on the raw probability UDs; bounded in
[0, 1], 1 meaning identical distributions.

```{r bhattacharyya}
bhatt <- function(p, q) {
  p <- terra::values(p); q <- terra::values(q)
  sum(sqrt(p * q), na.rm = TRUE)
}
bhatt(stk[[1]], stk[[2]])
```

Because both UDs already sum to 1 on the same grid, the returned
value is exactly the Bhattacharyya affinity used in the home-range
literature (e.g. Fieberg & Kochanny 2005).

### A visual: where the two UDs agree and disagree

```{r diff, fig.height = 5}
diff_ud <- stk[[1]] - stk[[2]]
terra::plot(diff_ud,
            main = "F3 UD - M3 UD",
            col = hcl.colors(100, "Blue-Red 3"))
```

Positive cells belong more to F3; negative to M3. Neutral grey
indicates the shared core.

## Earth-mover's distance (EMD)

EMD is the minimum-cost work to transport probability mass between
two UDs; it decreases smoothly as the two UDs drift into coincidence.
`emd()` computes pairwise EMDs over a stack of UDs and returns a
`stats::dist` object.

```{r emd}
d <- emd(stk)
d
```

By default `emd()` uses **Sinkhorn entropic regularisation**
(Cuturi 2013). On a small fisher stack the default call
completes in well under a minute. A `mask_quantile = 0.999`
pre-mask (via `ud_volume()`) drops cells outside the 99.9 %
volume contour — typically > 95 % of cells on well-supported UDs —
before transport, and is the single biggest practical speed-up.

For validation you can switch to the exact LP solver (requires the
`emdist` package):

```{r emd-exact, eval = FALSE}
d_exact <- emd(stk, method = "exact")

## relative difference between Sinkhorn and exact, typically sub-1%
max(abs(as.vector(d) - as.vector(d_exact)) / as.vector(d_exact))
```

### Unprojected (lon/lat) UDs and a hard distance cutoff

`emd()` defaults to Euclidean distance between cell centroids,
which is the right ground metric on projected UDs. If your UDs are
on a longitude/latitude grid (e.g. straight from `terra::rast()`
without `mt_aeqd_crs()`), pass `gc = TRUE` to use Haversine
great-circle distances instead — `emd()` warns when input is on a
longlat CRS and `gc = FALSE`. The `threshold` argument (in map
units when `gc = FALSE`, metres when `gc = TRUE`) clips the cost
matrix at a maximum transport distance, reproducing the EMD-hat
variant exposed by `move::emd()`. Both `gc` and `threshold` are
currently honoured by `method = "sinkhorn"` only; `method =
"exact"` is restricted to Euclidean, unclipped costs through
`emdist::emd()`.

### Tuning `reg` and `mask_quantile`

| What for | `reg` | `mask_quantile` | Expected behaviour |
|---|---|---|---|
| Default (balanced) | 0.05 | 0.999 | Fast; within \eqn{\sim}{~}0.1 % of exact |
| Fastest (interactive) | 0.1 | 0.95 | Very fast; within \eqn{\sim}{~}1 % |
| Highest fidelity | 0.01 | 0.9999 | Slow; near-exact |
| Exact | — | 0.999 | Use `method = "exact"` |

On irregularly supported or multi-modal UDs, consider raising
`mask_quantile` slightly so that secondary modes are not truncated.

## Pairwise comparison across a UD stack

Any metric that takes two UDs generalises to a distance matrix on a
stack. For a small number of UDs, a nested `sapply` is fine:

```{r pairwise, eval = FALSE}
ud_list <- lapply(seq_len(terra::nlyr(stk)), function(i) stk[[i]])
nm      <- names(stk)

mat <- outer(seq_along(ud_list), seq_along(ud_list),
             Vectorize(function(i, j) bhatt(ud_list[[i]], ud_list[[j]])))
dimnames(mat) <- list(nm, nm)
mat
```

For larger stacks the `emdist` EMD is the bottleneck (it is
super-linear in the number of non-zero cells), so downsample the
grid or mask to a tight contour first.

## Further reading

- `vignette("UD_dbbmm_ud", package = "move2utils")` — the Dynamic Brownian 
bridge UDs (dBBMM).
- `vignette("UD_dbgb_ud", package = "move2utils")` — the Directional UDs with 
the dynamic bivariate Gaussian bridge (dBGB).
- `vignette("UD_bursted_uds", package = "move2utils")` — producing dBBMM and 
dBGB UDs per behavioural or temporal segment per individual.
- `vignette("UD_gap_aware_ud", package = "move2utils")` — excluding long-gap 
segments from dBBMM/dBGB
- Fieberg, J., & Kochanny, C. O. (2005). Quantifying home-range
  overlap: the importance of the utilization distribution. *Journal
  of Wildlife Management*, 69(4), 1346–1359.
